---
title: "translocation_manu_16S_taxa"
author: "Songjun"
date: "2025-02-03"
output: html_document
---

#load package
```{r load backage} 
library(tidyverse)
library(gridExtra)
library(ggplot2)
library(reshape2)
library(ggrepel)
library(vegan)
library(viridis)
library(cowplot)
library(phyloseq)
library(SRS)
library(ANCOMBC)
library(ggpubr)
library(zCompositions)
library(readxl)
library(ggpubr)
library(emmeans)
library(dplyr)
library(kableExtra)
library(lme4)
library(lmerTest)
library(tidyr)
library(ggeffects)
library(purrr)
library(gllvm)



#many different colours
library(RColorBrewer)
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

```

#load data
#load resident cores data
```{r load data}

sample_meta_data_res <- read.csv("~/Desktop/Linnaeus University/OKG /T1 experiment analysis/16S_T1_ori_core/metadata.csv", sep=";")
sample_meta_data_res$Depth<-as.character(sample_meta_data_res$Depth)
sample_meta_data_res$Depth<-factor(sample_meta_data_res$Depth,levels = c("0","1","8","15","22","29"))
sample_meta_data_res$Core<-paste("Resident")
sample_meta_data_res$Time<-paste("T1")
sample_meta_data_res$Bay<-gsub('warm','heated',sample_meta_data_res$Bay)
sample_meta_data_res <- sample_meta_data_res %>%
  mutate(Core_ID = paste(Site, Core, sep = "_"))


feature_table_res<- read_tsv("~/Desktop/Linnaeus University/OKG /T1 experiment analysis/16S_T1_ori_core/ASV_table.tsv")

taxonomy_res<- read_tsv("~/Desktop/Linnaeus University/OKG /T1 experiment analysis/16S_T1_ori_core/ASV_tax_species.sbdi-gtdb.tsv")
#replace NA to 'unidentified'
taxonomy_res[is.na(taxonomy_res)]<-"unidentified"

asvs_res<-feature_table_res%>%
  gather(sample, count, 2:ncol(.)) %>% 
  filter(count > 0) %>%
  group_by(sample) %>% 
  mutate(relab = count/sum(count)) %>% 
  filter(n() > 10) %>% 
  ungroup()

feature_table_res%>%
  gather(sample, count, -1)%>%
  group_by(ASV_ID)%>%
  #filter(count > 0) %>% 
  ungroup() %>%
  mutate(count = as.integer(count))%>%
  # Add metadata
 inner_join(sample_meta_data_res, by = 'sample')%>%
  # Relative abundance, grouped first by sample name, and then next by rep(sampling replicate identifier)
  group_by(sample) %>%
  #group_by(rep)%>% #do at sample level first 
  mutate(relab = count/sum(count)) %>% ungroup() -> seqtab_res

```

#load translocated cores data
```{r load data}

sample_meta_data_trans <- read.csv("~/Desktop/Linnaeus University/OKG /T1 experiment analysis/16S_T1_trans_core/metadata.csv", sep=";")
sample_meta_data_trans$Depth<-as.character(sample_meta_data_trans$Depth)
sample_meta_data_trans$Depth<-factor(sample_meta_data_trans$Depth,levels = c("0","1","8","15","22","29"))
sample_meta_data_trans$Time<-paste("T1")
sample_meta_data_trans$Core<-paste("Translocated")
sample_meta_data_trans$Bay<-gsub('warm','heated',sample_meta_data_trans$Bay)
sample_meta_data_trans <- sample_meta_data_trans %>%
  mutate(Core_ID = paste(Site, Core, sep = "_"))

feature_table_trans<- read_tsv("~/Desktop/Linnaeus University/OKG /T1 experiment analysis/16S_T1_trans_core/ASV_table.tsv")

taxonomy_trans<- read_tsv("~/Desktop/Linnaeus University/OKG /T1 experiment analysis/16S_T1_trans_core/ASV_tax_species.sbdi-gtdb.tsv")
#replace NA to 'unidentified'
taxonomy_trans[is.na(taxonomy_trans)]<-"unidentified"

asvs_trans<-feature_table_trans%>%
  gather(sample, count, 2:ncol(.)) %>% 
  filter(count > 0) %>%
  group_by(sample) %>% 
  mutate(relab = count/sum(count)) %>% 
  filter(n() > 10) %>% 
  ungroup()

feature_table_trans%>%
  gather(sample, count, -1)%>%
  group_by(ASV_ID)%>%
  #filter(count > 0) %>% 
  ungroup() %>%
  mutate(count = as.integer(count))%>%
  # Add metadata
 inner_join(sample_meta_data_trans, by = 'sample')%>%
  # Relative abundance, grouped first by sample name, and then next by rep(sampling replicate identifier)
  group_by(sample) %>%
  #group_by(rep)%>% #do at sample level first 
  mutate(relab = count/sum(count)) %>% ungroup() -> seqtab_trans

```

#load T0 cores data
```{r}
sample_meta_data_t0 <- read.csv("~/Desktop/Linnaeus University/OKG /T0 experiment analysis/16S_2021_august_sediment/metadata.csv", sep=";")
sample_meta_data_t0$Depth<-as.character(sample_meta_data_t0$Depth)
sample_meta_data_t0$Depth<-factor(sample_meta_data_t0$Depth,levels = c("0","1","8","15","22","29"))
sample_meta_data_t0$Core<-paste("Resident")
sample_meta_data_t0$Time<-paste("T0")
sample_meta_data_t0$Bay<-gsub('warm','heated',sample_meta_data_t0$Bay)
sample_meta_data_t0 <- sample_meta_data_t0 %>%
  mutate(Core_ID = paste(Site, Core, sep = "_"))

feature_table_t0<- read_tsv("~/Desktop/Linnaeus University/OKG /T0 experiment analysis/16S_2021_august_sediment/ASV_table.tsv")

taxonomy_t0<- read_tsv("~/Desktop/Linnaeus University/OKG /T0 experiment analysis/16S_2021_august_sediment/ASV_tax_species.sbdi-gtdb.tsv")
#replace NA to 'unidentified'
taxonomy_t0[is.na(taxonomy_t0)]<-"unidentified"

asvs_t0<-feature_table_t0%>%
  gather(sample, count, 2:ncol(.)) %>% 
  filter(count > 0) %>%
  group_by(sample) %>% 
  mutate(relab = count/sum(count)) %>% 
  filter(n() > 10) %>% 
  ungroup()

feature_table_t0%>%
  gather(sample, count, -1)%>%
  group_by(ASV_ID)%>%
  #filter(count > 0) %>% 
  ungroup() %>%
  mutate(count = as.integer(count))%>%
  # Add metadata
 inner_join(sample_meta_data_t0, by = 'sample')%>%
  # Relative abundance, grouped first by sample name, and then next by rep(sampling replicate identifier)
  group_by(sample) %>%
  #group_by(rep)%>% #do at sample level first 
  mutate(relab = count/sum(count)) %>% ungroup() -> seqtab_t0

sum(asvs_t0$count)
sum(asvs_res$count)
sum(asvs_trans$count)

```

#t0 preparation
```{r}

## sample metadata
samples_df_T0 <- sample_meta_data_t0



## OTUs data
otu_mat_T0<- feature_table_t0

### tax data 

tax_mat_T0<- taxonomy_t0[,1:10]



### objects need to have row names 
otu_mat_T0 <- otu_mat_T0 %>%
    tibble::column_to_rownames("ASV_ID") 

tax_mat_T0 <- tax_mat_T0 %>% 
    tibble::column_to_rownames("ASV_ID")

 samples_df_T0 <- samples_df_T0 %>% 
    tibble::column_to_rownames("sample") 
 
 ###convert tax and otu to matrices 
otu_mat_T0 <- as.matrix(otu_mat_T0)
tax_mat_T0 <- as.matrix(tax_mat_T0)

## data is T_otu_mat
#transpose otu matrix so that sites are rows and species are in columns
T_otu_mat_T0<-t(otu_mat_T0)

trans_otu_mat_T0<-T_otu_mat_T0

trans_otu_mat_T0<-data.frame(sample= row.names(trans_otu_mat_T0), trans_otu_mat_T0)

```

# t1 preparation
```{r}

## sample metadata
samples_df_T1 <- rbind(sample_meta_data_res , sample_meta_data_trans)



## OTUs data
otu_mat_T1<- merge(feature_table_res,feature_table_trans,by = "ASV_ID",
                                   all = TRUE)
#replace NA to 0
otu_mat_T1[is.na(otu_mat_T1)]<-0

### tax data 

tax_mat_T1<- merge(taxonomy_res[,1:10],taxonomy_trans[,1:10],all = TRUE)
#remove duplicate
Dup_tax<-tax_mat_T1[duplicated(tax_mat_T1$ASV_ID),]
tax_mat_T1<-tax_mat_T1[-as.numeric(row.names.data.frame(Dup_tax)),]
rownames(tax_mat_T1)<-NULL


### objects need to have row names 
otu_mat_T1 <- otu_mat_T1 %>%
    tibble::column_to_rownames("ASV_ID") 

tax_mat_T1 <- tax_mat_T1 %>% 
    tibble::column_to_rownames("ASV_ID")

 samples_df_T1 <- samples_df_T1 %>% 
    tibble::column_to_rownames("sample") 
 
 ###convert tax and otu to matrices 
otu_mat_T1 <- as.matrix(otu_mat_T1)
tax_mat_T1 <- as.matrix(tax_mat_T1)

## data is T_otu_mat_T1
#transpose otu matrix so that sites are rows and species are in columns
T_otu_mat_T1<-t(otu_mat_T1)

trans_otu_mat_T1<-T_otu_mat_T1

trans_otu_mat_T1<-data.frame(sample= row.names(trans_otu_mat_T1), trans_otu_mat_T1)

```


#rarefaction curve for Suppplemental
```{r}


#rare_curve_t0
asvs_t0%>%
 dplyr::select(ASV_ID, sample, count) %>%
  spread(ASV_ID, count, fill = 0) %>%
  column_to_rownames("sample")%>%
rarecurve(sample= 100, step = 100, xlab = "Sequencing Depth", ylab = "# of ASVs",cex=0.4)

#rare_curve_t1_resident_core
asvs_res%>%
 dplyr::select(ASV_ID, sample, count) %>%
  spread(ASV_ID, count, fill = 0) %>%
  column_to_rownames("sample")%>%
rarecurve(sample= 100, step = 100, xlab = "Sequencing Depth", ylab = "# of ASVs",cex=0.4)

#rare_curve_t1_translocated_core
asvs_trans%>%
 dplyr::select(ASV_ID, sample, count) %>%
  spread(ASV_ID, count, fill = 0) %>%
  column_to_rownames("sample")%>%
rarecurve(sample= 100, step = 100, xlab = "Sequencing Depth", ylab = "# of ASVs",cex=0.4)



```



#mix t1 data together
```{r}

asvs_t1<-rbind(asvs_res,asvs_trans)

taxonomy_t1<-bind_rows(taxonomy_res[,1:9],taxonomy_trans[,1:9])%>% distinct(ASV_ID, .keep_all = TRUE)



sample_meta_data_t1<-rbind(sample_meta_data_res,sample_meta_data_trans)

```




#relatieve abundance bar plot(family level)

#select top 15 families for all data
```{r}

taxonomy_all<-bind_rows(taxonomy_t0[,1:9],taxonomy_t1[,1:9])%>% distinct(ASV_ID, .keep_all = TRUE)
sample_meta_data_all<-rbind(sample_meta_data_t0,sample_meta_data_t1)

p_all<- rbind(asvs_t0,asvs_t1) %>%
  inner_join(taxonomy_all, by = 'ASV_ID') %>%
  inner_join(sample_meta_data_all, by = 'sample')%>%
  group_by(Family, sample) %>%
  # Calculate the relative abundance of each family in each sample
  summarise(relab = sum(relab)) %>%
  # Calculate the *mean* relative abundance of each family over the samples
  summarise(mean_relab = sum(relab)) %>%
  ungroup() %>%
  # Filter out non-assigned family
  filter(!is.na(Family))%>%
  # Select the top 15
  top_n(16, mean_relab)


p_all

p_all_without_unidentified<-p_all[-16,]

```



#t0
```{r}


#depth
depth_Family_t0<-asvs_t0[,-4]%>% 
  filter(count > 0)%>%
  # Add metadata
 inner_join(sample_meta_data_t0, by = 'sample')%>%
  inner_join(taxonomy_t0, by = 'ASV_ID')%>%
  group_by(Bay, Depth)%>%
  # Calculate the relative abundance of each family in each site
  mutate(relab = count/sum(count))%>%
  ungroup()%>% 
  left_join(
    p_all %>% 
      transmute(Family, top_15_Family = Family),
    by = 'Family'
  ) %>%
  replace_na(list('top_15_Family' = 'Other_Families'))

ra_plot_table_t0<-depth_Family_t0%>%
  group_by(Bay, Depth,top_15_Family)%>%
  summarise(relab=sum(relab))


ra_plot_table_t0_only_top15<-ra_plot_table_t0[ra_plot_table_t0$top_15_Family != 'Other_Families',  ]
ra_plot_table_t0_only_top15<-ra_plot_table_t0_only_top15[ra_plot_table_t0_only_top15$top_15_Family !=  'unidentified', ]


ra_plot_t0_all<-ggplot(ra_plot_table_t0,aes(x=Depth, y=relab, fill=top_15_Family)) +
facet_grid(~Bay)+
geom_col() +
theme_bw() +
scale_fill_manual(values = col_vector)+
scale_x_discrete(limits=rev)+
coord_flip()+
labs(x="Sediment depth (cm)", y=element_blank())+
  theme(legend.position="bottom",
    legend.key.size = unit(0.5, 'cm'), 
        legend.key.height = unit(0.5, 'cm'), 
        legend.key.width = unit(0.7, 'cm'), 
        legend.title = element_blank(), 
        legend.text = element_text(size=10),
    axis.text.x = element_blank(),
    axis.ticks.x = element_line()) 

ra_plot_t0_all


ra_plot_t0_seleted<-ggplot(ra_plot_table_t0_only_top15,aes(x=Depth, y=relab, fill=top_15_Family)) +
facet_grid(~Bay)+
geom_col() +
theme_bw() +
scale_fill_manual(values = col_vector)+
scale_x_discrete(limits=rev)+
coord_flip()+
ggtitle("T0")+
labs(x="Sediment depth (cm)", y=element_blank())+
  theme(legend.position="bottom",
    legend.key.size = unit(0.5, 'cm'), 
        legend.key.height = unit(0.5, 'cm'), 
        legend.key.width = unit(0.7, 'cm'), 
        #legend.title = element_blank(), 
        legend.text = element_text(size=10),
    axis.ticks.x = element_line()) 

ra_plot_t0_seleted

```

#t1

#t1 heated bay
```{r}


#depth
depth_Family_t1<-asvs_t1[,-4]%>% 
  filter(count > 0)%>%
  # Add metadata
 inner_join(sample_meta_data_t1, by = 'sample')%>%
  inner_join(taxonomy_t1, by = 'ASV_ID')%>%
  group_by(Bay, Core, Depth)%>%
  # Calculate the relative abundance of each family in each site
  mutate(relab = count/sum(count))%>%
  ungroup()%>% 
  left_join(
    p_all %>% 
      transmute(Family, top_15_Family = Family),
    by = 'Family'
  ) %>%
  replace_na(list('top_15_Family' = 'Other_Families'))

ra_plot_table_t1_heated<-depth_Family_t1%>%
  filter(Bay == "heated")%>%
  group_by(Core, Depth,top_15_Family)%>%
  summarise(relab=sum(relab))

  
ra_plot_t1_heated_all<-ggplot(ra_plot_table_t1_heated,aes(x=Depth, y=relab, fill=top_15_Family)) +
facet_grid(~Core)+
geom_col() +
theme_bw() +
scale_fill_manual(values = col_vector)+
scale_x_discrete(limits=rev)+
coord_flip()+
labs(x="Sediment depth (cm)", y=element_blank())+
  theme(legend.position="bottom",
    legend.key.size = unit(0.5, 'cm'), 
        legend.key.height = unit(0.5, 'cm'), 
        legend.key.width = unit(0.7, 'cm'), 
        legend.title = element_blank(), 
        legend.text = element_text(size=10),
    axis.ticks.x = element_line(),  
        axis.text.x = element_blank()) 

ra_plot_t1_heated_all



ra_plot_table_t1_heated_only_top15<-ra_plot_table_t1_heated[ra_plot_table_t1_heated$top_15_Family != 'Other_Families',  ]
ra_plot_table_t1_heated_only_top15<-ra_plot_table_t1_heated_only_top15[ra_plot_table_t1_heated_only_top15$top_15_Family !=  'unidentified', ]



ra_plot_t1_heated_seleted<-ggplot(ra_plot_table_t1_heated_only_top15,aes(x=Depth, y=relab, fill=top_15_Family)) +
facet_grid(~Core)+
geom_col() +
theme_bw() +
scale_fill_manual(values = col_vector)+
scale_x_discrete(limits=rev)+
coord_flip()+
ggtitle("T1 heated bay")+
labs(x="Sediment depth (cm)", y=element_blank())+
  theme(legend.position="bottom",
    legend.key.size = unit(0.5, 'cm'), 
        legend.key.height = unit(0.5, 'cm'), 
        legend.key.width = unit(0.7, 'cm'), 
        #legend.title = element_blank(), 
        legend.text = element_text(size=10),
    axis.ticks.x = element_line()) 

ra_plot_t1_heated_seleted


```

#t1 control bay
```{r}

#depth
depth_Family_t1<-asvs_t1[,-4]%>% 
  filter(count > 0)%>%
  # Add metadata
 inner_join(sample_meta_data_t1, by = 'sample')%>%
  inner_join(taxonomy_t1, by = 'ASV_ID')%>%
  group_by(Bay, Core, Depth)%>%
  # Calculate the relative abundance of each family in each site
  mutate(relab = count/sum(count))%>%
  ungroup()%>% 
  left_join(
    p_all%>% 
      transmute(Family, top_15_Family = Family),
    by = 'Family'
  ) %>%
  replace_na(list('top_15_Family' = 'Other_Families'))

ra_plot_table_t1_control<-depth_Family_t1%>%
  filter(Bay == "control")%>%
  group_by(Core, Depth,top_15_Family)%>%
  summarise(relab=sum(relab))

  
ra_plot_t1_control_all<-ggplot(ra_plot_table_t1_control,aes(x=Depth, y=relab, fill=top_15_Family)) +
facet_grid(~Core)+
geom_col() +
theme_bw() +
scale_fill_manual(values = col_vector)+
scale_x_discrete(limits=rev)+
coord_flip()+
labs(x="Sediment depth (cm)", y=element_blank())+
  theme(legend.position="bottom",
    legend.key.size = unit(0.5, 'cm'), 
        legend.key.height = unit(0.5, 'cm'), 
        legend.key.width = unit(0.7, 'cm'), 
        legend.title = element_blank(), 
        legend.text = element_text(size=10),
    axis.ticks.x = element_line(),  
        axis.text.x = element_blank()) 

ra_plot_t1_control_all


ra_plot_table_t1_control_only_top15<-ra_plot_table_t1_control[ra_plot_table_t1_control$top_15_Family != 'Other_Families',  ]
ra_plot_table_t1_control_only_top15<-ra_plot_table_t1_control_only_top15[ra_plot_table_t1_control_only_top15$top_15_Family !=  'unidentified', ]


ra_plot_t1_control_seleted<-ggplot(ra_plot_table_t1_control_only_top15,aes(x=Depth, y=relab, fill=top_15_Family)) +
facet_grid(~Core)+
geom_col() +
theme_bw() +
scale_fill_manual(values = col_vector)+
scale_x_discrete(limits=rev)+
coord_flip()+
ggtitle("T1 control bay")+
labs(x="Sediment depth (cm)", y=element_blank())+
  theme(legend.position="bottom",
    legend.key.size = unit(0.5, 'cm'), 
        legend.key.height = unit(0.5, 'cm'), 
        legend.key.width = unit(0.7, 'cm'), 
        #legend.title = element_blank(), 
        legend.text = element_text(size=10),
    axis.ticks.x = element_line()) 

ra_plot_t1_control_seleted

```


#try mixed barplots
```{r}

top15_mix<-ggarrange(
  ra_plot_t0_seleted,
  ra_plot_t1_control_seleted,
  ra_plot_t1_heated_seleted, 
  labels = c("A", "B","C"), 
  font.label = list(size = 18),
  nrow  = 3,
  common.legend = TRUE,
  legend = "bottom"
)

ggsave("top15_mix.pdf", plot = top15_mix,width = 40, height =60,units = "cm")

supp_all_mix<-ggarrange(
  ra_plot_t0_all,
  ra_plot_t1_control_all,
  ra_plot_t1_heated_all, 
  labels = c("A", "B","C"), 
  font.label = list(size = 18),
  nrow  = 3,
  common.legend = TRUE,
  legend = "bottom"
)

ggsave("supp_all_mix.pdf", plot = supp_all_mix,width = 40, height =60,units = "cm")



```




#relative abundance LMMs for the top families

#T0
```{r}


# Prepare phyloseq object
ps_T0 <- phyloseq(
  otu_table(otu_mat_T0, taxa_are_rows = TRUE),
  tax_table(tax_mat_T0),
  sample_data(samples_df_T0)
)

# Aggregate to family level
ps_T0_family <- tax_glom(ps_T0, taxrank = "Family")

# Select top 15 families
top_15_families <-p_all_without_unidentified$Family 



  
otu_count_family_T0 <- as.data.frame(otu_table(ps_T0_family)) %>%
  mutate(Family = as.vector(tax_table(ps_T0_family)[, "Family"]))

otu_count_family_T0_15 <- otu_count_family_T0 %>%
  mutate(Family = ifelse(Family %in% top_15_families, Family, "Other")) %>%
  group_by(Family) %>%
  summarise(across(where(is.numeric), sum), .groups = "drop")

otu_mat_T0_15 <- otu_count_family_T0_15 %>%
  column_to_rownames("Family") %>%
  t() %>%
  as.matrix()

# Prepare metadata
sample_meta_T0 <- sample_data(ps_T0_family) %>%
  data.frame() %>%
  mutate(
    Core_ID = factor(Core_ID),
    Bay = factor(Bay),
    Core = factor(Core),
    Depth = as.numeric(as.character(Depth))
  ) %>%
  filter(rownames(.) %in% rownames(otu_mat_T0_15))

# Validate matching
stopifnot(all(rownames(otu_mat_T0_15) == rownames(sample_meta_T0)))

# Define covariates and design
X_T0 <- sample_meta_T0[, c("Bay", "Depth")] %>%
  mutate(
    Bay = factor(Bay, levels = c("control", "heated")),
    Depth = as.numeric(scale(Depth))  # standardise
  )

options('contrasts')
options('contrasts' = c('contr.treatment', 'contr.poly'))

study_design_T0 <- data.frame(Core_ID = sample_meta_T0$Core_ID)

# Define offset
offset_vec_T0 <- log(rowSums(otu_mat_T0_15))

# Fit gllvm model
fit_T0_top15 <- gllvm(
  y = otu_mat_T0_15,
  X = X_T0,
  formula = ~ Bay * Depth,
  family = "negative.binomial",
  offset = offset_vec_T0,
  studyDesign = study_design_T0,
  row.eff = ~(1 | Core_ID),
  num.lv = 0
)
fit_T0_top15_1 <- gllvm(
  y = otu_mat_T0_15,
  X = X_T0,
  formula = ~ Bay * Depth,
  family = "negative.binomial",
  offset = offset_vec_T0,
  studyDesign = study_design_T0,
  row.eff = ~(1 | Core_ID),
  num.lv = 1
)



fit_T0_top15_2 <- gllvm(
  y = otu_mat_T0_15,
  X = X_T0,
  formula = ~ Bay * Depth,
  family = "negative.binomial",
  offset = offset_vec_T0,
  studyDesign = study_design_T0,
  row.eff = ~(1 | Core_ID),
  num.lv = 2
)

#use AIC to select number of latent variables
AIC(fit_T0_top15)
AIC(fit_T0_top15_1)# 1 or 2 are best fit
AIC(fit_T0_top15_2)

#residual analysis
par(mfrow = c(1,2))
plot(fit_T0_top15_1, which = 1:2)

#visualize species correlations
cr_T0 <- getResidualCor(fit_T0_top15_1)
library(corrplot); library(gclus)
corrplot(cr_T0[order.single(cr_T0), order.single(cr_T0)], diag = FALSE, type = "lower", 
         method = "square", tl.cex = 1, tl.srt = 45, tl.col = "red")
         
# Summarise and export
summary_gllvm_T0 <- summary(fit_T0_top15_1)
gllvm_results_T0 <- as.data.frame(summary_gllvm_T0$Coef.tableX)
write.csv(gllvm_results_T0, "gllvm_results_T0_top15.csv", row.names = TRUE)

# Plot latent space
gllvm::ordiplot(fit_T0_top15_1)
coefplot(fit_T0_top15_1)

# Plot effects per family
coefs_T0 <- coef(fit_T0_top15_1)$Xcoef
ses_T0 <- fit_T0_top15_1$sd$Xcoef
beta0_T0 <- coef(fit_T0_top15_1)$Intercept
taxa_names_T0 <- intersect(rownames(coefs_T0), rownames(ses_T0))

bay_levels_T0 <- levels(X_T0$Bay)
depth_seq_T0 <- seq(min(X_T0$Depth), max(X_T0$Depth), length.out = 100)

newdata_T0 <- expand.grid(Bay = bay_levels_T0, Depth = depth_seq_T0) %>%
  mutate(
    Bay = factor(Bay, levels = bay_levels_T0),
    Depth = as.numeric(Depth)
  )

Xmat_no_int_T0 <- model.matrix(~ Bay * Depth, data = newdata_T0)[, -1, drop = FALSE]

pred_list_T0 <- lapply(taxa_names_T0, function(fam) {
  beta <- as.numeric(coefs_T0[fam, ])
  beta_se <- as.numeric(ses_T0[fam, ])
  intercept <- beta0_T0[fam]
  
  if (any(is.na(beta)) || any(is.na(beta_se))) return(NULL)
  
  eta <- intercept + Xmat_no_int_T0 %*% beta
  se_eta <- sqrt(rowSums(sweep(Xmat_no_int_T0, 2, beta_se, `*`)^2))
  
  tibble(
    Bay = newdata_T0$Bay,
    Depth = newdata_T0$Depth,
    Family = fam,
    predicted = exp(eta),
    lower = exp(eta - 1.96 * se_eta),
    upper = exp(eta + 1.96 * se_eta)
  )
})

pred_df_T0 <- bind_rows(pred_list_T0)

# Plot
p_gllvm_T0_15 <- ggplot(pred_df_T0, aes(x = Depth, y = predicted, colour = Bay, fill = Bay)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_wrap(~ Family, scales = "free_y") +
  scale_colour_manual(values = c("control" = "blue", "heated" = "orange")) +
  scale_fill_manual(values = c("control" = "blue", "heated" = "orange")) +
  labs(x = "Depth", y = "Predicted abundance", colour = "Bay", fill = "Bay") +
  theme_minimal(base_size = 12)+
  theme(
    strip.text = element_text(size = 7)  # decrease size of title
  )

p_gllvm_T0_15

ggsave("p_gllvm_T0_top15.jpg", plot = p_gllvm_T0_15, width = 17, height = 12, units = "cm", dpi = 1200)


## coefficient plot in ggplot to modify
library(tidyverse)
# Reshape coefficients to long format
coef_T0_long <- as.data.frame(as.table(coefs_T0)) %>%
  rename(Family = Var1, Variable = Var2, Estimate = Freq) %>%
  mutate(
    SE = as.vector(ses_T0)
  )



# Calculate 95% CI and significance
coef_T0_long <- coef_T0_long %>%
  mutate(
    lower = Estimate - 1.96 * SE,
    upper = Estimate + 1.96 * SE,
    sig = ifelse(lower * upper > 0, "yes", "no")  # significant if CI does not cross 0
  )

# Clean variable names
coef_T0_long$Variable <- coef_T0_long$Variable %>%
  str_replace("Bayheated:Depth", "Bay × Depth") %>%
  str_replace("Bayheated", "Bay") %>%
  str_replace("Depth", "Depth")
# Set Variable as a factor with desired order
coef_T0_long$Variable <- factor(coef_T0_long$Variable, levels = c("Bay", "Depth", "Bay × Depth"))

# fix the family order

family_levels <- rev(sort(unique(as.character(coef_T0_long$Family))))
coef_T0_long <- coef_T0_long %>%
  mutate(Family = factor(Family, levels = family_levels))



p_coef_T0 <- ggplot(coef_T0_long, aes(x = Estimate, y = Family, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0.2,
                 position = position_dodge(width = 0.5)) +
  geom_point(shape = 4, size = 2, position = position_dodge(width = 0.5)) +
  facet_wrap(~ Variable,  ncol = 3) +
  scale_colour_manual(values = c("yes" = "black", "no" = "grey70")) +
  labs(x = "Estimate", y = NULL, colour = "Significance") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 9),
    strip.text = element_text(size = 11),
   axis.title.x =  element_blank(), 
    legend.position = "none",
    panel.spacing = unit(1, "lines")
  )

p_coef_T0

# Save
ggsave("p_coef_T0_by_variable.jpg", plot = p_coef_T0, width = 20, height = 16, units = "cm", dpi = 1200)




```

#T1
```{r gllvm T1} 
# Make phyloseq object
ps_T1 <- phyloseq(
  otu_table(otu_mat_T1, taxa_are_rows = TRUE),
  tax_table(tax_mat_T1),
  sample_data(samples_df_T1)
)

# Aggregate to family level
ps_T1_family <- tax_glom(ps_T1, taxrank = "Family")

# Select top 15 families
top_15_families <-p_all_without_unidentified$Family 

otu_count_family_T1 <- as.data.frame(otu_table(ps_T1_family)) %>%
  mutate(Family = as.vector(tax_table(ps_T1_family)[, "Family"]))

otu_count_family_T1_15 <- otu_count_family_T1 %>%
  mutate(Family = ifelse(Family %in% top_15_families, Family, "Other")) %>%
  group_by(Family) %>%
  summarise(across(where(is.numeric), sum), .groups = "drop")

otu_mat_T1_15 <- otu_count_family_T1_15 %>%
  column_to_rownames("Family") %>%
  t() %>%
  as.matrix()

# Prepare metadata
sample_meta_T1 <- sample_data(ps_T1_family) %>%
  data.frame() %>%
  mutate(
    Core_ID = factor(Core_ID),
    Bay = factor(Bay),
    Core = factor(Core),
    Depth = as.numeric(as.character(Depth))
  ) %>%
  filter(rownames(.) %in% rownames(otu_mat_T1_15))

# Validate matching
stopifnot(all(rownames(otu_mat_T1_15) == rownames(sample_meta_T1)))

# Define covariates and design
X_T1 <- sample_meta_T1[, c("Bay", "Core", "Depth")] %>%
  mutate(
    Bay = factor(Bay),
    Core = factor(Core),
    Depth = as.numeric(scale(Depth))  # standardise
  )

study_design_T1 <- data.frame(Core_ID = sample_meta_T1$Core_ID)

# Define offset
offset_vec_T1 <- log(rowSums(otu_mat_T1_15))

# Fit gllvm model
fit_T1_top15 <- gllvm(
  y = otu_mat_T1_15,
  X = X_T1,
  formula = ~ Bay * Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1,
  studyDesign = study_design_T1,
  row.eff = ~(1 | Core_ID),
  num.lv = 0
)
fit_T1_top15_1 <- gllvm(
  y = otu_mat_T1_15,
  X = X_T1,
  formula = ~ Bay * Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1,
  studyDesign = study_design_T1,
  row.eff = ~(1 | Core_ID),
  num.lv = 1
)
fit_T1_top15_2 <- gllvm(
  y = otu_mat_T1_15,
  X = X_T1,
  formula = ~ Bay * Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1,
  studyDesign = study_design_T1,
  row.eff = ~(1 | Core_ID),
  num.lv = 2
)

#use AIC to select number of latent variables
AIC(fit_T1_top15)
AIC(fit_T1_top15_1)# 1 or 2 are best fit
AIC(fit_T1_top15_2)

#residual analysis
par(mfrow = c(1,2))
plot(fit_T1_top15_1, which = 1:2)

#visualize species correlations
cr_T1 <- getResidualCor(fit_T1_top15_1)
library(corrplot); library(gclus)
corrplot(cr_T1[order.single(cr_T1), order.single(cr_T1)], diag = FALSE, type = "lower", 
         method = "square", tl.cex = 1, tl.srt = 45, tl.col = "red")
         
# Summarise and export
summary_gllvm_T1 <- summary(fit_T1_top15_1)
gllvm_results_T1 <- as.data.frame(summary_gllvm_T1$Coef.tableX)
write.csv(gllvm_results_T1, "gllvm_results_T1_top15.csv", row.names = TRUE)

# Plot latent space
gllvm::ordiplot(fit_T1_top15_1)

#coefficient plot!!
coefplot(fit_T1_top15_1, mfrow=c(2,4))

# Plot effects per family
coefs_T1 <- coef(fit_T1_top15_1)$Xcoef
ses_T1 <- fit_T1_top15_1$sd$Xcoef
beta0_T1 <- coef(fit_T1_top15_1)$Intercept
taxa_names_T1 <- intersect(rownames(coefs_T1), rownames(ses_T1))

bay_levels_T1 <- levels(X_T1$Bay)
core_levels_T1<- levels(X_T1$Core)
depth_seq_T1 <- seq(min(X_T1$Depth), max(X_T1$Depth), length.out = 100)

newdata_T1 <- expand.grid(Bay = bay_levels_T1, Core = core_levels_T1, Depth = depth_seq_T1) %>%
  mutate(
    Bay = factor(Bay, levels = bay_levels_T1),
    Core = factor(Core, levels = core_levels_T1),
    Depth = as.numeric(Depth)
  )

Xmat_no_int_T1 <- model.matrix(~ Bay * Core * Depth, data = newdata_T1)[, -1, drop = FALSE]

pred_list_T1 <- lapply(taxa_names_T1, function(fam) {
  beta <- as.numeric(coefs_T1[fam, ])
  beta_se <- as.numeric(ses_T1[fam, ])
  intercept <- beta0_T1[fam]
  
  if (any(is.na(beta)) || any(is.na(beta_se))) return(NULL)
  
  eta <- intercept + Xmat_no_int_T1 %*% beta
  se_eta <- sqrt(rowSums(sweep(Xmat_no_int_T1, 2, beta_se, `*`)^2))
  
  tibble(
    Bay = newdata_T1$Bay,
    Core = newdata_T1$Core,
    Depth = newdata_T1$Depth,
    Family = fam,
    predicted = exp(eta),
    lower = exp(eta - 1.96 * se_eta),
    upper = exp(eta + 1.96 * se_eta)
  )
})

pred_df_T1 <- bind_rows(pred_list_T1)

# Plot
library(ggh4x)

p_gllvm_T1_15 <- ggplot(pred_df_T1, aes(x = Depth, y = predicted, colour = Core, fill = Core)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_nested(Family ~ Bay, scales = "free_y", space = "free_x") +  
  scale_colour_manual(values = c("red4", "cyan4")) +
  scale_fill_manual(values = c("red4", "cyan4")) +
  labs(x = "Depth", y = "Predicted abundance") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 10),
    legend.position = "none"
  )

p_gllvm_T1_15
ggsave("p_gllvm_T1_top15.jpg", plot = p_gllvm_T1_15, width = 15, height = 50, units = "cm", dpi = 1200)

p_gllvm_T1_15_control <- ggplot(pred_df_T1[pred_df_T1$Bay=="control",], aes(x = Depth, y = predicted, colour = Core, fill = Core)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_wrap(~ Family, scales = "free_y") +
  scale_colour_manual(values = c("red4", "cyan4")) +
  scale_fill_manual(values = c("red4", "cyan4")) +
  labs(x = "Depth", y = "Predicted abundance") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 7)
)

p_gllvm_T1_15_control
ggsave("p_gllvm_T1_control_top15.jpg", plot = p_gllvm_T1_15_control, width = 17, height = 12, units = "cm", dpi = 1200)


p_gllvm_T1_15_heated <- ggplot(pred_df_T1[pred_df_T1$Bay=="heated",], aes(x = Depth, y = predicted, colour = Core, fill = Core)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_wrap(~ Family, scales = "free_y") +
  scale_colour_manual(values = c("red4", "cyan4")) +
  scale_fill_manual(values = c("red4", "cyan4")) +
  labs(x = "Depth", y = "Predicted abundance") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 7)
)

p_gllvm_T1_15_heated
ggsave("p_gllvm_T1_heated_top15.jpg", plot = p_gllvm_T1_15_heated, width = 17, height = 12, units = "cm", dpi = 1200)



```


```{r gllvm T1 bays separate} 
# Subset metadata and OTU table by Bay
sample_meta_T1_control <- sample_meta_T1 %>% filter(Bay == "control")
sample_meta_T1_heated <- sample_meta_T1 %>% filter(Bay == "heated")

otu_mat_T1_15_control <- otu_mat_T1_15[rownames(sample_meta_T1_control), ]
otu_mat_T1_15_heated <- otu_mat_T1_15[rownames(sample_meta_T1_heated), ]

# Prepare design matrices for each Bay
X_T1_control <- sample_meta_T1_control %>%
  mutate(
    Core = factor(Core),
    Depth = as.numeric(scale(Depth)) # standardise within Bay
  )

X_T1_heated <- sample_meta_T1_heated %>%
  mutate(
    Core = factor(Core),
    Depth = as.numeric(scale(Depth))
  )

# Study design
study_design_T1_control <- data.frame(Core_ID = sample_meta_T1_control$Core_ID)
study_design_T1_heated <- data.frame(Core_ID = sample_meta_T1_heated$Core_ID)

# Define offsets
offset_vec_T1_control <- log(rowSums(otu_mat_T1_15_control))
offset_vec_T1_heated <- log(rowSums(otu_mat_T1_15_heated))

# Fit models separately
fit_T1_control <- gllvm(
  y = otu_mat_T1_15_control,
  X = X_T1_control,
  formula = ~ Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1_control,
  studyDesign = study_design_T1_control,
  row.eff = ~(1 | Core_ID),
  num.lv = 1
)

fit_T1_heated <- gllvm(
  y = otu_mat_T1_15_heated,
  X = X_T1_heated,
  formula = ~ Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1_heated,
  studyDesign = study_design_T1_heated,
  row.eff = ~(1 | Core_ID),
  num.lv = 1
)

# Compare model AICs 
AIC(fit_T1_control)
AIC(fit_T1_heated)

# Summarise and export
summary_control <- summary(fit_T1_control)
summary_heated <- summary(fit_T1_heated)

write.csv(as.data.frame(summary_control$Coef.tableX), "gllvm_results_T1_control.csv", row.names = TRUE)
write.csv(as.data.frame(summary_heated$Coef.tableX), "gllvm_results_T1_heated.csv", row.names = TRUE)

# check plots of residuals
par(mfrow = c(1, 2))
plot(fit_T1_control, which = 1:2)
plot(fit_T1_heated, which = 1:2)

# latent space plots
gllvm::ordiplot(fit_T1_control, main = "Control Bay")
gllvm::ordiplot(fit_T1_heated, main = "Heated Bay")

#coefficient plots!!
coefplot(fit_T1_control)
coefplot(fit_T1_heated)

#prediction plots
# Define prediction grid for each Bay
core_levels_control <- levels(X_T1_control$Core)
core_levels_heated <- levels(X_T1_heated$Core)
depth_seq_control <- seq(min(X_T1_control$Depth), max(X_T1_control$Depth), length.out = 100)
depth_seq_heated <- seq(min(X_T1_heated$Depth), max(X_T1_heated$Depth), length.out = 100)
newdata_control <- expand.grid(Core = core_levels_control, Depth = depth_seq_control)
newdata_heated <- expand.grid(Core = core_levels_heated, Depth = depth_seq_heated)
Xmat_control <- model.matrix(~ Core * Depth, data = newdata_control)[, -1, drop = FALSE]
Xmat_heated <- model.matrix(~ Core * Depth, data = newdata_heated)[, -1, drop = FALSE]
# Extract coefficients
coefs_control <- coef(fit_T1_control)$Xcoef
ses_control <- fit_T1_control$sd$Xcoef
beta0_control <- coef(fit_T1_control)$Intercept
coefs_heated <- coef(fit_T1_heated)$Xcoef
ses_heated <- fit_T1_heated$sd$Xcoef
beta0_heated <- coef(fit_T1_heated)$Intercept
taxa_names_control <- intersect(rownames(coefs_control), rownames(ses_control))
taxa_names_heated <- intersect(rownames(coefs_heated), rownames(ses_heated))
# Predict function
predict_gllvm <- function(coefs, ses, intercepts, taxa_names, Xmat, newdata, Bay_label) {
  lapply(taxa_names, function(fam) {
    beta <- as.numeric(coefs[fam, ])
    beta_se <- as.numeric(ses[fam, ])
    intercept <- intercepts[fam]
    if (any(is.na(beta)) || any(is.na(beta_se))) return(NULL)
    eta <- intercept + Xmat %*% beta
    se_eta <- sqrt(rowSums(sweep(Xmat, 2, beta_se, `*`)^2))
    tibble(
      Core = newdata$Core,
      Depth = newdata$Depth,
      Family = fam,
      Bay = Bay_label,
      predicted = exp(eta),
      lower = exp(eta - 1.96 * se_eta),
      upper = exp(eta + 1.96 * se_eta)
    )
  }) %>% bind_rows()
}
# Predict
pred_df_control <- predict_gllvm(coefs_control, ses_control, beta0_control, taxa_names_control, Xmat_control, newdata_control, Bay_label = "control")
pred_df_heated <- predict_gllvm(coefs_heated, ses_heated, beta0_heated, taxa_names_heated, Xmat_heated, newdata_heated, Bay_label = "heated")
# Combine predictions
pred_df_T1_separate <- bind_rows(pred_df_control, pred_df_heated)

library(ggh4x)

p_gllvm_T1_separate <- ggplot(pred_df_T1_separate, aes(x = Depth, y = predicted, colour = Core, fill = Core)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_nested(Family ~ Bay, scales = "free_y", space = "free_x") +  
  scale_colour_manual(values = c("red4", "cyan4")) +
  scale_fill_manual(values = c("red4", "cyan4")) +
  labs(x = "Depth", y = "Predicted abundance") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 10),
    legend.position = "none"
  )

p_gllvm_T1_separate

ggsave("p_gllvm_T1_separate_top15.jpg", plot = p_gllvm_T1_separate, width = 15, height = 50, units = "cm", dpi = 1200)


#Combined Coefficient Plot
library(tidyverse)
coef_control_long <- as.data.frame(as.table(coefs_control)) %>%
  rename(Family = Var1, Variable = Var2, Estimate = Freq) %>%
  mutate(
    SE = as.vector(ses_control),
    Bay = "control"
  )
coef_heated_long <- as.data.frame(as.table(coefs_heated)) %>%
  rename(Family = Var1, Variable = Var2, Estimate = Freq) %>%
  mutate(
    SE = as.vector(ses_heated),
    Bay = "heated"
  )
# Combine
coef_long <- bind_rows(coef_control_long, coef_heated_long)
# Keep only predictors of interest
coef_long <- coef_long %>%
  filter(str_detect(Variable, "Core|Depth"))
# Calculate 95% CIs and significance
coef_long <- coef_long %>%
  mutate(
    lower = Estimate - 1.96 * SE,
    upper = Estimate + 1.96 * SE,
    sig = ifelse(lower * upper > 0, "yes", "no")  # significant if CI does not cross 0
  )
# Clean variable names
coef_long$Variable <- coef_long$Variable %>%
  str_replace("CoreTranslocated:Depth", "Treatment × Depth") %>%
  str_replace("CoreTranslocated", "Treatment") %>%
  str_replace("Depth", "Depth")
# Set Variable as a factor with desired order
coef_T0_long$Variable <- factor(coef_T0_long$Variable, levels = c("Treatment", "Depth", "Treatment × Depth"))

# Sort Families by effect size **within each Bay and Variable**
coef_long <- coef_long %>%
  group_by(Bay, Variable) %>%
  mutate(Family = fct_reorder(Family, Estimate)) %>%
  ungroup()


coef_long$Variable<-factor(coef_long$Variable, levels = c("Treatment","Depth", "Treatment × Depth") )

coef_long <- coef_long %>%
  mutate(Family = factor(Family, levels = family_levels))


p_coef_final <- ggplot(coef_long, aes(x = Estimate, y = Family, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0.2,
                 position = position_dodge(width = 0.5)) +
  geom_point(shape = 4, size = 2, position = position_dodge(width = 0.5)) +  # Cross
  facet_grid(Bay ~ Variable, space = "free") +
  scale_colour_manual(values = c("yes" = "black", "no" = "grey70")) +
  labs(x = "Estimate", y = NULL, colour = "Significance") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 9),
    strip.text = element_text(size = 11),
    legend.position = "none",
    panel.spacing = unit(1, "lines")
  )

p_coef_final

ggsave("p_coef_final_by_bay_variable_T1.jpg", plot = p_coef_final, width = 22, height = 16, units = "cm", dpi = 1200)

```


#try mix coef t0 t1 into one plot
```{r}

p_coef_mix<-ggarrange(
  p_coef_T0,
  p_coef_final,
  labels = c("A", "B"), 
  font.label = list(size = 18),
  nrow  = 2,
  heights  = c(1, 2)
)

p_coef_mix


ggsave("p_coef_mix.pdf", plot = p_coef_mix, width = 20, height = 25, units = "cm", dpi = 1200)

```


