---
title: "16S_translocation"
author: "Songjun, edited by Romana"
date: "2026-03-05"
output: html_document
---

#load package
```{r load backage} 
library(tidyverse)
library(gridExtra)
library(ggplot2)
library(reshape2)
library(ggrepel)
library(vegan)
library(viridis)
library(cowplot)
library(phyloseq)
library(SRS)
library(ANCOMBC)
library(ggpubr)
library(zCompositions)
library(readxl)
library(ggpubr)
library(emmeans)
library(DHARMa)
library(lmerTest)
library(ggeffects)
library(rstatix)

#new colours for Ida's thesis:
bay_exp_colors <- c(
  "Heated_Resident"      = "#D6191B",
  "Control_Resident"     = "#2166AC",
  "Heated_Translocated"  = "#00ACC1",
  "Control_Translocated" = "#F57C00"
)  

#many different colours
library(RColorBrewer)
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

```


#load data
#load resident cores data
```{r load data}

sample_meta_data_res <- read.csv("T1_res/metadata.csv", sep=";")
sample_meta_data_res$Depth<-as.character(sample_meta_data_res$Depth)
sample_meta_data_res$Depth<-factor(sample_meta_data_res$Depth,levels = c("0","1","8","15","22","29"))
sample_meta_data_res$Core<-paste("Resident")
sample_meta_data_res$Time<-paste("T1")
sample_meta_data_res$Bay<-gsub('warm','Heated',sample_meta_data_res$Bay)
sample_meta_data_res$Bay<-gsub('control','Control',sample_meta_data_res$Bay)
sample_meta_data_res <- sample_meta_data_res %>%
  mutate(Core_ID = paste(Site, Core, sep = "_"))


feature_table_res<- read_tsv("T1_res/ASV_table.tsv")

taxonomy_res<- read_tsv("T1_res/ASV_tax_species.sbdi-gtdb.tsv")
#replace NA to 'unidentified'
taxonomy_res[is.na(taxonomy_res)]<-"unidentified"

asvs_res<-feature_table_res%>%
  gather(sample, count, 2:ncol(.)) %>% 
  filter(count > 0) %>%
  group_by(sample) %>% 
  mutate(relab = count/sum(count)) %>% 
  filter(n() > 10) %>% 
  ungroup()

feature_table_res%>%
  gather(sample, count, -1)%>%
  group_by(ASV_ID)%>%
  #filter(count > 0) %>% 
  ungroup() %>%
  mutate(count = as.integer(count))%>%
  # Add metadata
 inner_join(sample_meta_data_res, by = 'sample')%>%
  # Relative abundance, grouped first by sample name, and then next by rep(sampling replicate identifier)
  group_by(sample) %>%
  #group_by(rep)%>% #do at sample level first 
  mutate(relab = count/sum(count)) %>% ungroup() -> seqtab_res

```

#load translocated cores data
```{r load data}

sample_meta_data_trans <- read.csv("T1_trans/metadata.csv", sep=";")
sample_meta_data_trans$Depth<-as.character(sample_meta_data_trans$Depth)
sample_meta_data_trans$Depth<-factor(sample_meta_data_trans$Depth,levels = c("0","1","8","15","22","29"))
sample_meta_data_trans$Time<-paste("T1")
sample_meta_data_trans$Core<-paste("Translocated")
sample_meta_data_trans$Bay<-gsub('warm','Heated',sample_meta_data_trans$Bay)
sample_meta_data_trans$Bay<-gsub('control','Control',sample_meta_data_trans$Bay)
sample_meta_data_trans <- sample_meta_data_trans %>%
  mutate(Core_ID = paste(Site, Core, sep = "_"))

feature_table_trans<- read_tsv("T1_trans/ASV_table.tsv")

taxonomy_trans<- read_tsv("T1_trans/ASV_tax_species.sbdi-gtdb.tsv")
#replace NA to 'unidentified'
taxonomy_trans[is.na(taxonomy_trans)]<-"unidentified"

asvs_trans<-feature_table_trans%>%
  gather(sample, count, 2:ncol(.)) %>% 
  filter(count > 0) %>%
  group_by(sample) %>% 
  mutate(relab = count/sum(count)) %>% 
  filter(n() > 10) %>% 
  ungroup()

feature_table_trans%>%
  gather(sample, count, -1)%>%
  group_by(ASV_ID)%>%
  #filter(count > 0) %>% 
  ungroup() %>%
  mutate(count = as.integer(count))%>%
  # Add metadata
 inner_join(sample_meta_data_trans, by = 'sample')%>%
  # Relative abundance, grouped first by sample name, and then next by rep(sampling replicate identifier)
  group_by(sample) %>%
  #group_by(rep)%>% #do at sample level first 
  mutate(relab = count/sum(count)) %>% ungroup() -> seqtab_trans

```

#load T0 cores data
```{r}
sample_meta_data_t0 <- read.csv("T0_data/metadata.csv", sep=";")
sample_meta_data_t0$Depth<-as.character(sample_meta_data_t0$Depth)
sample_meta_data_t0$Depth<-factor(sample_meta_data_t0$Depth,levels = c("0","1","8","15","22","29"))
sample_meta_data_t0$Core<-paste("Resident")
sample_meta_data_t0$Time<-paste("T0")
sample_meta_data_t0$Bay<-gsub('warm','Heated',sample_meta_data_t0$Bay)
sample_meta_data_t0$Bay<-gsub('control','Control',sample_meta_data_t0$Bay)
sample_meta_data_t0 <- sample_meta_data_t0 %>%
  mutate(Core_ID = paste(Site, Core, sep = "_"))

feature_table_t0<- read_tsv("T0_data/ASV_table.tsv")

taxonomy_t0<- read_tsv("T0_data/ASV_tax_species.sbdi-gtdb.tsv")
#replace NA to 'unidentified'
taxonomy_t0[is.na(taxonomy_t0)]<-"unidentified"

asvs_t0<-feature_table_t0%>%
  gather(sample, count, 2:ncol(.)) %>% 
  filter(count > 0) %>%
  group_by(sample) %>% 
  mutate(relab = count/sum(count)) %>% 
  filter(n() > 10) %>% 
  ungroup()

feature_table_t0%>%
  gather(sample, count, -1)%>%
  group_by(ASV_ID)%>%
  #filter(count > 0) %>% 
  ungroup() %>%
  mutate(count = as.integer(count))%>%
  # Add metadata
 inner_join(sample_meta_data_t0, by = 'sample')%>%
  # Relative abundance, grouped first by sample name, and then next by rep(sampling replicate identifier)
  group_by(sample) %>%
  #group_by(rep)%>% #do at sample level first 
  mutate(relab = count/sum(count)) %>% ungroup() -> seqtab_t0

sum(asvs_t0$count)
sum(asvs_res$count)
sum(asvs_trans$count)

```


#load translocation temperature
```{r Temperature}
Hobo_data_cages <- read.csv("Hobo_data_cages.csv",header = TRUE)
Hobo_data_cages$Date <- gsub( " .*$", "", Hobo_data_cages$Time)
Hobo_data_cages <- Hobo_data_cages %>%
  mutate(Date = as.Date(Date, format="%d/%m/%Y"))

# Filter the data to include Sediment and dates until february 2022
Hobo_data_sediment <- Hobo_data_cages %>%
  filter(Place == "Sediment" & Date <= as.Date("2022-02-28"))
head(Hobo_data_sediment)

# Calculate the average temperature for each site
average_temperature <- Hobo_data_sediment %>%
  group_by(Bay, Site) %>%
  summarize(Average_Temperature = mean(Temperature, na.rm = TRUE), .groups = "drop")

average_temperature$Bay<-gsub('Warm','Heated',average_temperature$Bay)

bay_avg <- average_temperature %>%
  group_by(Bay) %>%
  summarize(Average_Temperature = mean(Average_Temperature, na.rm = TRUE))

# add missing sites C and I and use average temperature values from each bay
missing_sites <- tibble(
  Bay = c("Heated", "Control"),
  Site = c("C", "I")
) %>%
  left_join(bay_avg, by = "Bay")
average_temperature_complete <- average_temperature %>%
  bind_rows(missing_sites)

```



#rarefaction curves
```{r}
#rare_curve_t0
asvs_t0%>%
 dplyr::select(ASV_ID, sample, count) %>%
  spread(ASV_ID, count, fill = 0) %>%
  column_to_rownames("sample")%>%
rarecurve(sample= 100, step = 100, xlab = "Sequencing Depth", ylab = "# of ASVs",cex=0.4,
            xlim = c(0, 250000),  
            ylim = c(0,5000))


#rare_curve_t1 resident cores
asvs_res%>%
 dplyr::select(ASV_ID, sample, count) %>%
  spread(ASV_ID, count, fill = 0) %>%
  column_to_rownames("sample")%>%
rarecurve(sample= 100, step = 100, xlab = "Sequencing Depth", ylab = "# of ASVs",cex=0.4,
            xlim = c(0, 250000),  
            ylim = c(0,5000))


#rare_curve_translocated cores
asvs_trans%>%
 dplyr::select(ASV_ID, sample, count) %>%
  spread(ASV_ID, count, fill = 0) %>%
  column_to_rownames("sample")%>%
rarecurve(sample= 100, step = 100, xlab = "Sequencing Depth", ylab = "# of ASVs",cex=0.4,
            xlim = c(0, 250000),  
            ylim = c(0,5000))

```


#alpha diversity

#t0

```{r Alpha Diversity Function - SRS - Preparation }

input_data_t0 <- asvs_t0%>%  
  dplyr::select(ASV_ID, sample, count)%>%  
  spread(sample,count, fill= 0) %>% 
  remove_rownames() %>% 
  column_to_rownames(var = "ASV_ID")

#(e.g. species counts of the library with the lowest sequencing depth):
Cmin <- min(colSums(input_data_t0))
Cmin

```

```{r Alpha diversity based on SRS}

#---------------running the SRS function on example data####
SRS_output_t0 <- SRS(data = input_data_t0, Cmin = Cmin)
SRS_output_t0


###Diversity Index Shannon
###between bays
SRSshannon_t0<- SRS_output_t0 %>%
  t()%>%
  data.frame()%>%
  rownames_to_column(var="sample")%>%
  plyr::ddply(~sample, function(x) {vegan::diversity(x[-1], index="shannon")}) %>%
  dplyr::rename(shannon= V1) %>% #Shannon 
  left_join(sample_meta_data_t0, by="sample")



```

#T0 Alpha diversity Statistics (Linear regression)
```{r}

SRSshannon_T0b <-SRSshannon_t0
SRSshannon_T0b$Depth <- as.numeric(as.character(SRSshannon_T0b$Depth))
T0_alpha<-lmer(shannon~Bay*Depth+(1|Core_ID), data = SRSshannon_T0b)
summary(T0_alpha)
anova(T0_alpha) # significant Core*Bay interaction 
rand(T0_alpha) #random effect sign
eta_squared(T0_alpha)

plot(T0_alpha)
hist(resid(T0_alpha))
simulateResiduals(fittedModel = T0_alpha, plot=TRUE)

emmeans(T0_alpha, spec="Bay",by="Depth", contr="pairwise")# significantly higher shannon diversity  in the resident cores of the heated than control bay
emmeans(T0_alpha, spec="Depth",by="Bay", contr="pairwise")# significantly higher shannon diversity  in the resident cores of the heated than control bay
emtrends(T0_alpha, ~ Bay, var = "Depth")

```


#T0 visualization
```{r }

#whole column box plot
shanSRS_T0_plot <- SRSshannon_t0 %>%
  ggplot(aes(x = Bay, y = shannon,fill=Bay)) + 
  geom_boxplot() +
  scale_fill_manual(values=c("#2166AC","#D6191B"))+
  labs(x = "", y = "Shannon's H Index") +
  coord_cartesian(ylim = c(4, 7.5)) +
  #ggtitle("T0 alpha diversity")+
  theme_bw() +
  theme(text = element_text(size = 20)) +
  theme(panel.grid = element_blank(), legend.position = "none")

shanSRS_T0_plot



T0_alpha_pred <- ggeffect(T0_alpha, terms = c("Depth", "Bay"))

T0_alpha_predictionplot <- ggplot(T0_alpha_pred, aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, colour = NA) +
  geom_point(data = SRSshannon_T0b,
             mapping = aes(x = Depth, y = shannon, color = Bay),
             position = position_jitter(width = 0.1), alpha = 0.5,
             inherit.aes = FALSE) +
  scale_colour_manual(values = c("#2166AC", "#D6191B"),name="Bay") +
  scale_fill_manual(values = c("#2166AC", "#D6191B"),name="Bay") +
  labs(x = "Depth (cm)", y = "Shannon's H Index") +
  coord_cartesian(ylim = c(4, 7.5)) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  theme(panel.grid = element_blank(), legend.position = "none")

T0_alpha_predictionplot
 
#mix plots together

T0_alpha_mixed_boxplot <- ggarrange(
  shanSRS_T0_plot,
  T0_alpha_predictionplot,  
  labels = c("a", "b"), 
  font.label = list(size = 18),
  ncol  = 2,
  widths = c(1, 1.5)
)

T0_alpha_mixed_boxplot

ggsave("T0_alpha_mixed.pdf", plot = T0_alpha_mixed_boxplot,width = 40, height = 20,units = "cm")




```





#t1

#shannon H values calculation
```{r Alpha Diversity Function - SRS - Preparation }

input_data_t1 <- rbind(asvs_res,asvs_trans)%>%  
  dplyr::select(ASV_ID, sample, count)%>%  
  spread(sample,count, fill= 0) %>% 
  remove_rownames() %>% 
  column_to_rownames(var = "ASV_ID")

#(e.g. species counts of the library with the lowest sequencing depth):
Cmin <- min(colSums(input_data_t1))
Cmin

```


```{r Alpha diversity based on SRS}

#---------------running the SRS function on example data####
SRS_output_t1 <- SRS(data = input_data_t1, Cmin = Cmin)
SRS_output_t1


sample_meta_data_t1<-rbind(sample_meta_data_res,sample_meta_data_trans)

###Diversity Index Shannon
SRSshannon_t1<- SRS_output_t1 %>%
  t()%>%
  data.frame()%>%
  rownames_to_column(var="sample")%>%
  plyr::ddply(~sample, function(x) {vegan::diversity(x[-1], index="shannon")}) %>%
  dplyr::rename(shannon= V1) %>% #Shannon 
  left_join(sample_meta_data_t1, by="sample")




```

#T1 Statistics (Linear model) 

```{r}

###Statistical approach - Alpha diversity
##Linear regression 
SRSshannon_t1b <-SRSshannon_t1
SRSshannon_t1b$Depth <- as.numeric(as.character(SRSshannon_t1b$Depth))
SRSshannon_t1b$Bay <- as.factor(SRSshannon_t1b$Bay)
SRSshannon_t1b$Core <- as.factor(SRSshannon_t1b$Core)

T1_alpha<-lmer(shannon~Bay*Depth*Core+(1|Core_ID), data = SRSshannon_t1b)

summary(T1_alpha)
anova(T1_alpha)
rand(T1_alpha)
eta_squared(T1_alpha)

plot(T1_alpha)
hist(resid(T1_alpha))
simulateResiduals(fittedModel = T1_alpha, plot=TRUE)


T1_alpha2<-lmer(shannon~Bay*Depth+Bay*Core+Depth*Core+(1|Core_ID), data = SRSshannon_t1b)

summary(T1_alpha2)
anova(T1_alpha2)
rand(T1_alpha2)
eta_squared(T1_alpha2)


```



#T1 visualization
```{r }

# Define colours
bay_exp_colors <- c(
  "Heated_Resident"      = "#D6191B", # #D6191B
  "Control_Resident"     = "#2166AC", # #2166AC
  "Heated_Translocated"  = "#00ACC1", # bluish green
  "Control_Translocated" = "#F57C00"  # reddish purple
)


bay_exp_labels <- c(
  "Heated_Resident"      = "Resident\nheated",
  "Control_Resident"     = "Resident\ncontrol",
  "Heated_Translocated"  = "Translocated\nC->H",
  "Control_Translocated" = "Translocated\nH->C"
)

# Add color_group to data
SRSshannon_t1$color_group  <- paste(SRSshannon_t1$Bay,  SRSshannon_t1$Core,  sep = "_")
SRSshannon_t1b$color_group <- paste(SRSshannon_t1b$Bay, SRSshannon_t1b$Core, sep = "_")


#whole column box plot

shannon_boxplot_t1<-ggplot(SRSshannon_t1, aes(x = color_group, y = shannon,fill=color_group)) + 
  geom_boxplot(linewidth = 0.75) +
  labs(x = "", y = "Shannon's H Index" ) +
  coord_cartesian(ylim = c(5, 7.5)) +
  scale_fill_manual(
  values = bay_exp_colors,
  labels = bay_exp_labels,
  name   = "Treatment"
)+
  scale_x_discrete(
  labels = bay_exp_labels
)+
  theme_bw() +
  theme(text = element_text(size = 16),
        axis.title = element_text(size = 16),  # axis titles (x and y labels)
        axis.text = element_text(size = 12),   # axis tick labels
        strip.text = element_text(size = 16),  # facet labels
        panel.border = element_rect(linewidth = 1.25),
        #panel.grid = element_blank(), 
        plot.margin = margin(5, 30, 5, 5), 
        legend.position = "none")

shannon_boxplot_t1



#raw dot plot with depths

T1_alpha_prediction <- predict_response(T1_alpha, terms = c("Depth", "Core", "Bay"))
names(T1_alpha_prediction)[names(T1_alpha_prediction) == "facet"] <- "Bay"

# Add color_group to prediction object (group = Core, Bay = facet column)
T1_alpha_prediction$color_group <- paste(T1_alpha_prediction$Bay, 
                                          T1_alpha_prediction$group, sep = "_")


T1_alpha_predictionplot <- ggplot(T1_alpha_prediction, aes(x = x, y = predicted, color = color_group, fill = color_group)) +
  geom_line(aes(linetype = group),size = 1.25) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, colour = NA) +
  geom_point(data = SRSshannon_t1b,
             mapping = aes(x = Depth, y = shannon, color = color_group),
             position = position_jitter(width = 0.1), alpha = 0.5,
             inherit.aes = FALSE) +
  scale_colour_manual(values = bay_exp_colors,name="Group") +
  scale_fill_manual(values = bay_exp_colors,name="Group") +
  labs(x = "Depth (cm)", y = "Shannon's H Index") +
  coord_cartesian(ylim = c(5, 7.5)) +
  facet_wrap(~Bay, nrow = 2, 
           labeller = labeller(Bay = c("Heated" = "Heated bay", 
                                       "Control" = "Control bay")))+
  theme_bw()+
  theme(text = element_text(size = 16), 
        axis.title = element_text(size = 16),  # axis titles (x and y labels)
        axis.text = element_text(size = 12),   # axis tick labels
        strip.text = element_text(size = 16),  # facet labels
        panel.border = element_rect(linewidth = 1.25),
        strip.background = element_rect(fill = "white", linewidth = 1.25),
        legend.position = "none")
 
T1_alpha_predictionplot

 
#mix plots together

T1_alpha_mixed_boxplot <- ggarrange(
  shannon_boxplot_t1,
  T1_alpha_predictionplot,  
  labels = c("a", "b"), 
  font.label = list(size = 18),
  ncol  = 2,
  widths = c(1, 1.5)
)

T1_alpha_mixed_boxplot

ggsave("T1_alpha_mixed.pdf", plot = T1_alpha_mixed_boxplot,width = 30, height = 18,units = "cm")
ggsave("T1_alpha_mixed.jpg", plot = T1_alpha_mixed_boxplot,width = 30, height = 18,units = "cm")



```

# T0 vs T1 Residents Alpha Diversity Comparison
```{r }
# Combine resident data from T0 and T1
# Filter T1 to residents only and add timepoint label
SRSshannon_t1_res <- SRSshannon_t1 %>%
  filter(Core == "Resident") %>%  # adjust to match your factor level name
  mutate(Timepoint = "T1")

SRSshannon_t0_res <- SRSshannon_t0 %>%
  mutate(Timepoint = "T0")

# Combine
SRSshannon_res_combined <- bind_rows(SRSshannon_t0_res, SRSshannon_t1_res)
SRSshannon_res_combined$Depth <- as.numeric(as.character(SRSshannon_res_combined$Depth))
SRSshannon_res_combined$Timepoint <- as.factor(SRSshannon_res_combined$Timepoint)
SRSshannon_res_combined$Bay <- as.factor(SRSshannon_res_combined$Bay)

# LMM
T0T1_res_alpha <- lmer(shannon ~ Timepoint * Bay * Depth + (1|Core_ID), 
                        data = SRSshannon_res_combined)

summary(T0T1_res_alpha)
anova(T0T1_res_alpha)
rand(T0T1_res_alpha)
eta_squared(T0T1_res_alpha)

plot(T0T1_res_alpha)
hist(resid(T0T1_res_alpha))
simulateResiduals(fittedModel = T0T1_res_alpha, plot = TRUE)

# Post-hoc
emmeans(T0T1_res_alpha, spec = "Timepoint", by = "Bay", contr = "pairwise")
emmeans(T0T1_res_alpha, spec = "Timepoint", by = "Depth", contr = "pairwise")
emtrends(T0T1_res_alpha, ~ Timepoint * Bay, var = "Depth")
```
#visualize with boxplot
```{r}
# Boxplot - Shannon by Timepoint and Bay
shannon_boxplot_T0T1_res <- ggplot(SRSshannon_res_combined, 
                                    aes(x = Timepoint, y = shannon, fill = Bay)) + 
  geom_boxplot(linewidth = 0.75) +
  scale_fill_manual(values = c("#2166AC", "#D6191B")) +
  labs(x = "", y = "Shannon's H Index") +
  coord_cartesian(ylim = c(4, 7.5)) +
  theme_bw() +
  theme(text = element_text(size = 16))  +
  theme(text = element_text(size = 16), 
        axis.title = element_text(size = 16),  # axis titles (x and y labels)
        axis.text = element_text(size = 12),   # axis tick labels
        strip.text = element_text(size = 16),  # facet labels
        panel.border = element_rect(linewidth = 1.25),
        strip.background = element_rect(linewidth = 1.25),
        plot.margin = margin(5, 30, 5, 5),
        legend.position = c(0.85, 0.15),
        legend.background = element_rect(fill = "white", 
                                         colour = "grey80", linewidth = 0.5))

shannon_boxplot_T0T1_res

# Prediction plot
T0T1_res_pred <- ggeffect(T0T1_res_alpha, terms = c("Depth", "Timepoint", "Bay"))
names(T0T1_res_pred)[names(T0T1_res_pred) == "facet"] <- "Bay"

T0T1_res_pred <- ggeffect(T0T1_res_alpha, terms = c("Depth", "Bay", "Timepoint"))
names(T0T1_res_pred)[names(T0T1_res_pred) == "facet"] <- "Timepoint"

T0T1_res_predictionplot <- ggplot(T0T1_res_pred, 
                                   aes(x = x, y = predicted, color = group, fill = group)) +
  geom_line(size = 1.25) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, colour = NA) +
  geom_point(data = SRSshannon_res_combined,
             mapping = aes(x = Depth, y = shannon, color = Bay),
             position = position_jitter(width = 0.1), alpha = 0.5,
             inherit.aes = FALSE) +
  scale_colour_manual(values = c("#2166AC", "#D6191B"), name = "Bay") +
  scale_fill_manual(values = c("#2166AC", "#D6191B"), name = "Bay") +
  labs(x = "Depth (cm)", y = "Shannon's H Index") +
  coord_cartesian(ylim = c(4, 7.5)) +
  facet_wrap(~Timepoint, nrow = 2) +
  theme_bw() +
  theme(text = element_text(size = 16), 
        axis.title = element_text(size = 16),  # axis titles (x and y labels)
        axis.text = element_text(size = 12),   # axis tick labels
        strip.text = element_text(size = 16),  # facet labels
        panel.border = element_rect(linewidth = 1.25),
        strip.background = element_rect(fill = "white", linewidth = 1.25),
        legend.position = "none")

T0T1_res_predictionplot


# Combined plot
T0T1_res_mixed <- ggarrange(
  shannon_boxplot_T0T1_res,
  T0T1_res_predictionplot,
  labels = c("a", "b"),
  font.label = list(size = 16),
  ncol = 2,
  widths = c(1, 1.5)
)

T0T1_res_mixed

ggsave("T0T1_res_alpha_mixed.pdf", plot = T0T1_res_mixed, width = 30, height = 18, units = "cm")
ggsave("T0T1_res_alpha_mixed.jpg", plot = T0T1_res_mixed, width = 30, height = 18, units = "cm")
```


#beta diversity (PCoA)

#t0 preparation
```{r}

## sample metadata
samples_df_T0 <- sample_meta_data_t0



## OTUs data
otu_mat_T0<- feature_table_t0

### tax data 

tax_mat_T0<- taxonomy_t0[,1:10]



### objects need to have row names 
otu_mat_T0 <- otu_mat_T0 %>%
    tibble::column_to_rownames("ASV_ID") 

tax_mat_T0 <- tax_mat_T0 %>% 
    tibble::column_to_rownames("ASV_ID")

 samples_df_T0 <- samples_df_T0 %>% 
    tibble::column_to_rownames("sample") 
 
 ###convert tax and otu to matrices 
otu_mat_T0 <- as.matrix(otu_mat_T0)
tax_mat_T0 <- as.matrix(tax_mat_T0)

## data is T_otu_mat
#transpose otu matrix so that sites are rows and species are in columns
T_otu_mat_T0<-t(otu_mat_T0)

trans_otu_mat_T0<-T_otu_mat_T0

trans_otu_mat_T0<-data.frame(sample= row.names(trans_otu_mat_T0), trans_otu_mat_T0)

```

### calculated aitchison distance matrix
```{r}
# Robust CLR transformation (handles zeros, no pseudocount needed)
rclr_T0 <- decostand(T_otu_mat_T0, method = "rclr")

# Aitchison distance = Euclidean distance on rclr-transformed data
aitch_dist_T0 <- vegdist(rclr_T0, method = "euclidean")

```


###PCOA
```{r}
# calculate principal coordinates analysis (Bray-Curtis)
pcoa.aitch_dist_T0 <- cmdscale(aitch_dist_T0, k = 2, eig = T)

# extract axis positions for each site from cmdscale object and create a dataframe for plotting
pcoa.aitch_dist.plotting_T0 <- as.data.frame(pcoa.aitch_dist_T0$points)
colnames(pcoa.aitch_dist.plotting_T0) <- c("axis_1", "axis_2")


pcoa.aitch_dist.plotting_T0<-cbind(pcoa.aitch_dist.plotting_T0, samples_df_T0)

# calculate the proportion of variance in the data which is explained by the first two PCoA axes
pcoa.aitch_dist_T0$eig[1]/(sum(pcoa.aitch_dist_T0$eig))
# resulting value is the Xlab 
#0.2006358,  20.06% -> new value 0.5482268, 54.82%

pcoa.aitch_dist_T0$eig[2]/(sum(pcoa.aitch_dist_T0$eig))
#resulting value is the ylab 
#0.1434748,  14.35% -> new value 0.390013, 39.00%

```


#pcoa plot T0
```{r}
# create a PCoA plot

pcoa.aitch_dist.plotting_T0$Bay <- gsub('warm','heated',pcoa.aitch_dist.plotting_T0$Bay )


pcoa_plot_t0 <- ggplot(pcoa.aitch_dist.plotting_T0, aes(x = axis_1, y = axis_2)) +
  geom_point(aes(colour =Bay ,shape = Depth), size= 3.5) +
  scale_color_manual(values=c( "#2166AC","#D6191B"))+
  theme_bw() + 
  theme(text = element_text(size = 20)) +
  xlab("PC1 (54.8%)") + #change these as required
  ylab("PC2 (39.0%)")  #change these as required
  

pcoa_plot_t0

ggsave("pcoa_T0.pdf", plot = pcoa_plot_t0, width = 25, height = 17,units = "cm")


```

#following PERMANOVA for the pairwise distance matrix
```{r}

#make depth into continues numeric factor then make model
str(samples_df_T0)
samples_df_depth_numeric_T0<-samples_df_T0
samples_df_depth_numeric_T0$Depth<-as.numeric(samples_df_depth_numeric_T0$Depth)
str(samples_df_depth_numeric_T0)

adonis2(aitch_dist_T0~Bay*Depth+Bay/Site, data = samples_df_depth_numeric_T0)

library(permute)
perm <- how(
  blocks = samples_df_depth_numeric_T0$Core_ID,       # restrict permutations within cores
  within = Within(type = "free")     # freely permute within each core
)
perm <- with(samples_df_depth_numeric_T0, how(nperm = 999, blocks = Core_ID))
adonis2(aitch_dist_T0 ~ Bay*Depth, data = samples_df_depth_numeric_T0, permutations = perm , by="term")

# Betadisper - test homogeneity of dispersions (assumption check for PERMANOVA)
# Bay
disp_bay_T0 <- betadisper(aitch_dist_T0, samples_df_depth_numeric_T0$Bay)
anova(disp_bay_T0)
permutest(disp_bay_T0, permutations = 999)
plot(disp_bay_T0, main = "T0 - Bay")
boxplot(disp_bay_T0, main = "T0 - Bay")
#not sign

# Depth (as categorical)
disp_depth_T0 <- betadisper(aitch_dist_T0, samples_df_T0$Depth)
anova(disp_depth_T0)
permutest(disp_depth_T0, permutations = 999)
plot(disp_depth_T0, main = "T0 - Depth")
boxplot(disp_depth_T0, main = "T0 - Depth")
#not sign
```




# t1 preparation
```{r}

## sample metadata
samples_df_T1 <- rbind(sample_meta_data_res , sample_meta_data_trans)



## OTUs data
otu_mat_T1<- merge(feature_table_res,feature_table_trans,by = "ASV_ID",
                                   all = TRUE)
#replace NA to 0
otu_mat_T1[is.na(otu_mat_T1)]<-0

### tax data 

tax_mat_T1<- merge(taxonomy_res[,1:10],taxonomy_trans[,1:10],all = TRUE)
#remove duplicate
Dup_tax<-tax_mat_T1[duplicated(tax_mat_T1$ASV_ID),]
tax_mat_T1<-tax_mat_T1[-as.numeric(row.names.data.frame(Dup_tax)),]
rownames(tax_mat_T1)<-NULL


### objects need to have row names 
otu_mat_T1 <- otu_mat_T1 %>%
    tibble::column_to_rownames("ASV_ID") 

tax_mat_T1 <- tax_mat_T1 %>% 
    tibble::column_to_rownames("ASV_ID")

 samples_df_T1 <- samples_df_T1 %>% 
    tibble::column_to_rownames("sample") 
 
 ###convert tax and otu to matrices 
otu_mat_T1 <- as.matrix(otu_mat_T1)
tax_mat_T1 <- as.matrix(tax_mat_T1)

## data is T_otu_mat_T1
#transpose otu matrix so that sites are rows and species are in columns
T_otu_mat_T1<-t(otu_mat_T1)

trans_otu_mat_T1<-T_otu_mat_T1

trans_otu_mat_T1<-data.frame(sample= row.names(trans_otu_mat_T1), trans_otu_mat_T1)

```

### compare pairwise dissimilarity, all sites, aitchison
```{r}
rclr_T1 <- decostand(T_otu_mat_T1, method = "rclr")
aitch_dist_T1 <- vegdist(rclr_T1, method = "euclidean")
```

###PCOA
```{r}
# calculate principal coordinates analysis (Bray-Curtis)
pcoa.aitch_dist_T1 <- cmdscale(aitch_dist_T1, k = 2, eig = T)

# extract axis positions for each site from cmdscale object and create a dataframe for plotting
pcoa.aitch_dist_T1.plotting <- as.data.frame(pcoa.aitch_dist_T1$points)
colnames(pcoa.aitch_dist_T1.plotting) <- c("axis_1", "axis_2")

pcoa.aitch_dist_T1.plotting<-cbind(pcoa.aitch_dist_T1.plotting, samples_df_T1)

# calculate the proportion of variance in the data which is explained by the first two PCoA axes
pcoa.aitch_dist_T1$eig[1]/(sum(pcoa.aitch_dist_T1$eig))
# resulting value is the Xlab 
# 0.1886253,  18.86% -> new value 0.5200812, 52.01%

pcoa.aitch_dist_T1$eig[2]/(sum(pcoa.aitch_dist_T1$eig))
#resulting value is the ylab 
# 0.1193483,  11.93% -> new value 0.3902701, 39.03%


```


#pcoa plot T0 vs resident T1

#select resident core, and compared with T0 to check "bottle effect"
```{r}
# Combine T0 and T1 resident OTU matrices
# T1 resident sample names
res_samples_T1 <- rownames(samples_df_T1[samples_df_T1$Core == "Resident", ])

# Subset T1 to residents only
T_otu_mat_T1_res <- T_otu_mat_T1[res_samples_T1, ]

# Combine OTU matrices - fill missing ASVs with 0
T_otu_mat_bottle <- bind_rows(
  as.data.frame(T_otu_mat_T0),
  as.data.frame(T_otu_mat_T1_res)
) %>% replace(is.na(.), 0)

# Combined metadata
samples_df_T0_tp <- samples_df_T0 %>% mutate(Timepoint = "T0")
samples_df_T1_res <- samples_df_T1[res_samples_T1, ] %>% mutate(Timepoint = "T1")
samples_df_bottle <- bind_rows(samples_df_T0_tp, samples_df_T1_res)
samples_df_bottle$Depth <- as.factor(samples_df_bottle$Depth)

# rclr and Aitchison distance
rclr_bottle <- decostand(T_otu_mat_bottle, method = "rclr")
aitch_dist_bottle <- vegdist(rclr_bottle, method = "euclidean")

# PCoA
pcoa_bottle <- cmdscale(aitch_dist_bottle, k = 2, eig = TRUE)
pcoa_bottle_plotting <- as.data.frame(pcoa_bottle$points)
colnames(pcoa_bottle_plotting) <- c("axis_1", "axis_2")
pcoa_bottle_plotting <- cbind(pcoa_bottle_plotting, samples_df_bottle)

# Variance explained
pcoa_bottle$eig[1] / sum(pcoa_bottle$eig)
pcoa_bottle$eig[2] / sum(pcoa_bottle$eig)

# Plot - split into T0 and T1 panels, colour by Bay, shape by Depth
pcoa_bottle_T0 <- ggplot(pcoa_bottle_plotting[pcoa_bottle_plotting$Timepoint == "T0", ], 
                          aes(x = axis_1, y = axis_2)) +
  geom_point(aes(colour = Bay, shape = Depth), size = 3.5) +
  scale_color_manual(values = c("Control" = "#2166AC", "Heated" = "#D6191B"),name = "Bay:") +
  scale_shape_manual(values = c(16, 17, 15, 3, 7, 8),name = "Depth:" ) +
  theme_bw() +
  theme(text = element_text(size = 16),
        title = element_text(size = 16),
        axis.title = element_text(size = 16),  # axis titles (x and y labels)
        axis.text = element_text(size = 13),   # axis tick labels
        strip.text = element_text(size = 16),  # facet labels
        panel.border = element_rect(linewidth = 1.25),
        strip.background = element_rect(linewidth = 1.25),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 15),
        legend.position = "bottom", 
        legend.box = "horizontal",
        legend.spacing.x = unit(80, "pt"), 
        legend.margin = margin(15, 0, 0, 35),
        plot.margin = margin(5, 30, 5, 5)) +
  guides(colour = guide_legend(order = 1, nrow = 1),
  shape  = guide_legend(order = 2, nrow = 1)) +
  xlab("PC1 (49.6%)") +
  ylab("PC2 (41.6%)") +
  ggtitle("T0") +
  xlim(range(pcoa_bottle_plotting$axis_1)) +  # same axis limits for both panels
  ylim(range(pcoa_bottle_plotting$axis_2))

pcoa_bottle_T1 <- ggplot(pcoa_bottle_plotting[pcoa_bottle_plotting$Timepoint == "T1", ], 
                          aes(x = axis_1, y = axis_2)) +
  geom_point(aes(colour = Bay, shape = Depth), size = 3.5) +
  scale_color_manual(values = c("Control" = "#2166AC", "Heated" = "#D6191B"),name = "Bay:") +
  scale_shape_manual(values = c(16, 17, 15, 3, 7, 8),name = "Depth:" ) +
  theme_bw() +
  theme(text = element_text(size = 16),
        title = element_text(size = 16),
        axis.title = element_text(size = 16),  # axis titles (x and y labels)
        axis.text = element_text(size = 13),   # axis tick labels
        strip.text = element_text(size = 16),  # facet labels
        panel.border = element_rect(linewidth = 1.25),
        strip.background = element_rect(linewidth = 1.25),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 15),
        legend.position = "bottom", 
        legend.box = "horizontal",
        legend.spacing.x = unit(80, "pt"), 
        legend.margin = margin(15, 0, 0, 35)) +
  guides(colour = guide_legend(order = 1, nrow = 1),
  shape  = guide_legend(order = 2, nrow = 1)) +
  xlab("PC1 (49.6%)") +
  ylab("PC2 (41.6%)") +
  ggtitle("T1") +
  xlim(range(pcoa_bottle_plotting$axis_1)) +
  ylim(range(pcoa_bottle_plotting$axis_2))


# Combine with shared legend
bottle_effect_plot <- ggarrange(
  pcoa_bottle_T0,
  pcoa_bottle_T1,
  labels = c("a", "b"),
  font.label = list(size = 18, face="plain"),
  common.legend = TRUE,
  legend = "bottom",
  ncol = 2,
  align = "hv"
)
bottle_effect_plot
ggsave("bottle_effect_combined_pcoa.pdf", plot = bottle_effect_plot, width = 30, height = 15.8, units = "cm")

# PERMANOVA - significant Timepoint = bottle/time effect
samples_df_bottle$Depth <- as.numeric(as.character(samples_df_bottle$Depth))
perm_bottle <- with(samples_df_bottle, how(nperm = 999, blocks = Core_ID))
adonis2(aitch_dist_bottle ~ Timepoint * Bay * Depth,
        data = samples_df_bottle,
        permutations = perm_bottle,
        by = "term") #all significant

# Betadisper
disp_timepoint <- betadisper(aitch_dist_bottle, samples_df_bottle$Timepoint)
permutest(disp_timepoint, permutations = 999)#sign
boxplot(disp_timepoint, main = "Bottle effect - Timepoint")

```


#pcoa plot T1

#select resident core, and compared with T0 to check "bottle effect"
```{r}

pcoa_plot_t1_res<- ggplot(pcoa.aitch_dist_T1.plotting[pcoa.aitch_dist_T1.plotting$Core=="Resident",], aes(x = axis_1, y = axis_2)) +
  geom_point(aes(colour =Bay , shape=Depth), size= 3.5) +
  scale_color_manual(values=c("#2166AC","#D6191B"))+
  theme_bw() + 
  theme(text = element_text(size = 20)) +
  xlab("PC1 (52.0%)") + #change these as required
  ylab("PC2 (39.0%)") +
  #ggtitle("")+ 
  theme()

pcoa_plot_t1_res

bottle_effect_plot <- ggarrange(
  pcoa_plot_t0,  
  pcoa_plot_t1_res,  
  labels = c("a", "b"), 
  font.label = list(size = 18),
  common.legend = TRUE, 
  legend = "right", 
  nrow =   2
)

bottle_effect_plot

ggsave("bottle_effect(T0_pcoa+T1_res_pcoa).pdf", plot = bottle_effect_plot, width = 18, height = 30,units = "cm")

```


#separated by bay
```{r }
pcoa.aitch_dist_T1.plotting$color_group <- paste(
  pcoa.aitch_dist_T1.plotting$Bay,
  pcoa.aitch_dist_T1.plotting$Core, sep = "_")

bay_exp_colors <- c(
  "Heated_Resident"      = "#D6191B",
  "Control_Resident"     = "#2166AC",
  "Heated_Translocated"  = "#00ACC1",
  "Control_Translocated" = "#F57C00"
)

bay_exp_labels <- c(
  "Heated_Resident"      = "Resident heated",
  "Control_Resident"     = "Resident control",
  "Heated_Translocated"  = "Translocated C->H",
  "Control_Translocated" = "Translocated H->C"
)

# Control bay
pcoa_plot_t1_control <- ggplot(pcoa.aitch_dist_T1.plotting[pcoa.aitch_dist_T1.plotting$Bay=="Control",], 
                                aes(x = axis_1, y = axis_2)) +
  geom_point(aes(colour = color_group, shape = Depth), size = 3.5) +
  scale_color_manual(values = bay_exp_colors,  labels = bay_exp_labels, 
  name = "Treatment:", limits = names(bay_exp_colors))+
  theme_bw() + 
  ggtitle("Control bay") +
  xlab("PC1 (52.0%)") + 
  ylab("PC2 (39.0%)") +
  theme(text = element_text(size = 16),
        title = element_text(size = 16),
        axis.title = element_text(size = 16), 
        axis.text = element_text(size = 13),   
        strip.text = element_text(size = 16), 
        panel.border = element_rect(linewidth = 1.25),
        strip.background = element_rect(linewidth = 1.25),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 15),
        legend.position = "bottom", 
        legend.box = "vertical",
        legend.spacing.x = unit(80, "pt"), 
        legend.margin = margin(15, 0, 0, 35)) +
  guides(colour = guide_legend(order = 1, nrow = 2),
         shape  = guide_legend(order = 2, nrow = 1)) +
  xlim(range(pcoa.aitch_dist_T1.plotting$axis_1)) + 
  ylim(range(pcoa.aitch_dist_T1.plotting$axis_2))

# Heated bay
pcoa_plot_t1_heated <- ggplot(pcoa.aitch_dist_T1.plotting[pcoa.aitch_dist_T1.plotting$Bay=="Heated",], 
                               aes(x = axis_1, y = axis_2)) +
  geom_point(aes(colour = color_group, shape = Depth), size = 3.5) +
  scale_color_manual(values = bay_exp_colors,  labels = bay_exp_labels, 
  name = "Treatment:", limits = names(bay_exp_colors))+
  theme_bw() + 
  ggtitle("Heated bay") +
  xlab("PC1 (52.0%)") + 
  ylab("PC2 (39.0%)") +
  theme(text = element_text(size = 16),
        title = element_text(size = 16),
        axis.title = element_text(size = 16), 
        axis.text = element_text(size = 13),   
        strip.text = element_text(size = 16), 
        panel.border = element_rect(linewidth = 1.25),
        strip.background = element_rect(linewidth = 1.25),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 15),
        legend.position = "bottom", 
        legend.box = "vertical",
        legend.spacing.x = unit(80, "pt"), 
        legend.margin = margin(15, 0, 0, 35),
        plot.margin = margin(5, 30, 5, 5)) +
  guides(colour = guide_legend(order = 1, nrow = 2),
         shape  = guide_legend(order = 2, nrow = 1)) +
  xlim(range(pcoa.aitch_dist_T1.plotting$axis_1)) + 
  ylim(range(pcoa.aitch_dist_T1.plotting$axis_2))

pcoa_plot_t1_mixed <- ggarrange(
  pcoa_plot_t1_heated,  
  pcoa_plot_t1_control,  
  labels = c("a", "b"), 
  font.label = list(size = 18, face = "plain"),
  common.legend = TRUE, 
  legend = "bottom", 
  ncol = 2,
  align = "hv"
)

pcoa_plot_t1_mixed
ggsave("pcoa_T1_mixed.pdf", plot = pcoa_plot_t1_mixed, width = 30, height = 18, units = "cm")
```


#following PERMANOVA for the pairwise distance matrix
```{r}

#make depth into continues numeric factor then make model
str(samples_df_T1)
samples_df_depth_numeric_T1<-samples_df_T1
samples_df_depth_numeric_T1$Depth<-as.numeric(samples_df_depth_numeric_T1$Depth)
str(samples_df_depth_numeric_T1)


library(permute)

perm <- with(samples_df_depth_numeric_T1, how(nperm = 999, blocks = Core_ID))
adonis2(aitch_dist_T1 ~ Bay*Depth*Core, data = samples_df_depth_numeric_T1, permutations = perm , by="term")

# Betadisper - test homogeneity of dispersions (assumption check for PERMANOVA)
# Bay
disp_bay_T1 <- betadisper(aitch_dist_T1, samples_df_depth_numeric_T1$Bay)
permutest(disp_bay_T1, permutations = 999)
plot(disp_bay_T1, main = "T1 - Bay")
boxplot(disp_bay_T1, main = "T1 - Bay")#sign heated greater than control
disp_bay_T1$group.distances
# Depth (as categorical)
disp_depth_T1 <- betadisper(aitch_dist_T1, samples_df_T1$Depth)
permutest(disp_depth_T1, permutations = 999)
plot(disp_depth_T1, main = "T1 - Depth")
boxplot(disp_depth_T1, main = "T1 - Depth")# sign highest at 8 and 15
disp_depth_T1$group.distances
# Core 
disp_core_T1 <- betadisper(aitch_dist_T1, samples_df_T1$Core)
permutest(disp_core_T1, permutations = 999)
plot(disp_core_T1, main = "T1 - Core")
boxplot(disp_core_T1, main = "T1 - Core")#not sign
disp_core_T1$group.distances


#separate permanova for each bay

# control bay
samples_control <- samples_df_depth_numeric_T1 %>% filter(Bay == "Control")
attr(aitch_dist_T1, "Labels") <- rownames(samples_df_depth_numeric_T1)
aitch_control <- as.dist(as.matrix(aitch_dist_T1)[rownames(samples_control), rownames(samples_control)])
perm_control <- with(samples_control, how(within = Within(type = "free"), blocks = Core_ID, nperm = 999))
adonis_control <- adonis2(aitch_control ~ Depth*Core, data = samples_control, permutations = perm_control, by = "term")
adonis_control
# Betadisper - test homogeneity of dispersions (assumption check for PERMANOVA)
# Depth (as categorical)
disp_depth_control <- betadisper(aitch_control, samples_control$Depth)
permutest(disp_depth_control, permutations = 999)
boxplot(disp_depth_control, main = "control - Depth")# sign- highest at 8 and 15 cm, lowest 0 and 1 cm
disp_depth_control$group.distances
# Core 
disp_core_control <- betadisper(aitch_control, samples_control$Core)
permutest(disp_core_control, permutations = 999)
boxplot(disp_core_control, main = "control - Core")# higher in translocated
disp_core_control$group.distances
# heated bay
samples_heated <- samples_df_depth_numeric_T1 %>% filter(Bay == "Heated")
aitch_heated <- as.dist(as.matrix(aitch_dist_T1)[rownames(samples_heated), rownames(samples_heated)])
perm_heated <- with(samples_heated, how(within = Within(type = "free"), blocks = Core_ID, nperm = 999))
adonis_heated <- adonis2(aitch_heated ~ Depth*Core, data = samples_heated, permutations = perm_heated, by = "term")
adonis_heated
# Betadisper - test homogeneity of dispersions (assumption check for PERMANOVA)
# Depth (as categorical)
disp_depth_heated <- betadisper(aitch_heated, samples_heated$Depth)
permutest(disp_depth_heated, permutations = 999)
boxplot(disp_depth_heated, main = "heated - Depth")# sign- highest at 8 and 15 cm, lowest 0 and 1 cm
disp_depth_heated$group.distances
# Core 
disp_core_heated <- betadisper(aitch_heated, samples_heated$Core)
permutest(disp_core_heated, permutations = 999)
boxplot(disp_core_heated, main = "heated - Core")#ns

```


# RDA
```{r RDA T1 }
samples_df_T1$sample <- rownames(samples_df_T1)
#add temperature data to sample_df
average_temperature_complete$Bay<-gsub('Heated','heated',average_temperature_complete$Bay)
average_temperature_complete$Bay<-gsub('Control','control',average_temperature_complete$Bay)
samples_df_temp <- samples_df_T1 %>%
  left_join(average_temperature_complete, by = c("Bay", "Site"))

#add geochemistry
tube_T1 <- read.csv("geochem_tube_T1.csv")
tube_T1$Depth <- as.factor(tube_T1$Depth)
tube_T1$Bay<-gsub('Warm','heated',tube_T1$Bay)
tube_T1$Bay<-gsub('Control','control',tube_T1$Bay)
tube_T1$Core<-gsub('ori','Resident',tube_T1$Core)
tube_T1$Core<-gsub('trans','Translocated',tube_T1$Core)

samples_df_all <- samples_df_temp %>%
  left_join(tube_T1, by = c("Bay", "Site", "Depth", "Core"))

samples_df_all2 <- samples_df_all
samples_df_all2$Depth <- as.numeric(as.character(samples_df_all2$Depth))

#RDA with aitchison
rda_t1 <- rda(rclr_T1 ~ Average_Temperature+Depth+NO3..mg.L.+SO4..mg.L.+Total.Fe..μM.+PO4.3...μM.+organic._percentage, data = samples_df_all2, na.action = na.omit)
summary(rda_t1, display = NULL)#contribution to the mean squared Euclidean distance: RDA1: 40.5% , RDA2: 12.1%
vif.rda_t1 <- vif.cca(rda_t1) 
vif.rda_t1#all less than 2
#rda_t1_table <- anova.cca(rda_t1, by = "margin", permutations=999)
#rda_t1_table <- anova(rda_t1, by = "margin", permutations=999)
#rda_t1_table
#Extract site data 
scrs_rda_t1 <- scores(rda_t1, display=c("sp","wa","lc","bp","cn"))
df_sites_rda_t1<-data.frame(scrs_rda_t1$sites)
df_sites_rda_t1$sample <- samples_df_all$sample
df_sites_rda_t1 <- df_sites_rda_t1 %>% inner_join(samples_df_all, by="sample")
df_sites_rda_t1$Depth <- factor(as.character(df_sites_rda_t1$Depth), levels = c("0", "1", "8", "15", "22", "29"))
# Add color_group to RDA site scores
df_sites_rda_t1$color_group <- paste(df_sites_rda_t1$Bay, df_sites_rda_t1$Core, sep = "_")

#Draw biplots
df_arrows_rda_t1<- scrs_rda_t1$biplot*8
df_arrows_rda_t1=as.data.frame(df_arrows_rda_t1)
#edit the row names with significant stars directly/ remove non significant variables from the plot
df_arrows_rda_t1 <- subset(df_arrows_rda_t1, !rownames(df_arrows_rda_t1) %in% c("NO3..mg.L."))
df_arrows_rda_t1 <- subset(df_arrows_rda_t1, !rownames(df_arrows_rda_t1) %in% c("SO4..mg.L."))
df_arrows_rda_t1 <- subset(df_arrows_rda_t1, !rownames(df_arrows_rda_t1) %in% c("Total.Fe..μM."))
rownames(df_arrows_rda_t1) <- c("Temperature", "Depth","PO4","Organic C")

bay_exp_labels2 <- c(
  "heated_Resident"      = "Resident\nheated",
  "control_Resident"     = "Resident\ncontrol",
  "heated_Translocated"  = "Translocated\nC->H",
  "control_Translocated" = "Translocated\nH->C"
)

bay_exp_colors2 <- c(
  "heated_Resident"      = "#D6191B",
  "control_Resident"     = "#2166AC",
  "heated_Translocated"  = "#00ACC1",
  "control_Translocated" = "#F57C00"
)  


p_rda_t1 <- ggplot(data = df_sites_rda_t1, aes(RDA1, RDA2, colour = color_group)) +
  geom_point(aes(shape = Depth), size = 3) +
  scale_color_manual(values = bay_exp_colors2, labels = bay_exp_labels2,
                     name = "Treatment:", limits = names(bay_exp_colors2)) +
  #scale_shape_manual(values = c("Resident" = 16, "Translocated" = 8)) +
  #scale_alpha(range = c(0.3, 1)) +
  theme_bw() +
  xlim(-8, 8) +
  ylim(-8, 8) +
  labs(x = "RDA1 (40.5 %)", y = "RDA2 (12.1 %)") +
  geom_segment(data = df_arrows_rda_t1 * 0.9, 
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.2, "cm")), inherit.aes = FALSE) +
  geom_text(data = as.data.frame(df_arrows_rda_t1),
            aes(RDA1, RDA2, label = rownames(df_arrows_rda_t1)),
            inherit.aes = FALSE)+
  theme(text = element_text(size = 16),
        title = element_text(size = 16),
        axis.title = element_text(size = 16), 
        axis.text = element_text(size = 13),   
        strip.text = element_text(size = 16), 
        panel.border = element_rect(linewidth = 1.25),
        strip.background = element_rect(linewidth = 1.25),
        #legend.text = element_text(size = 14),
        #legend.title = element_text(size = 15),
        legend.position = "right", 
        legend.box = "vertical",
        #legend.spacing.x = unit(80, "pt"), 
        #legend.margin = margin(15, 0, 0, 35),
        #plot.margin = margin(5, 30, 5, 5)
        ) 
p_rda_t1



ggsave("p_rda_t1.pdf", plot = p_rda_t1, width = 18, height = 13,units = "cm",dpi=1200)



#rda_t1_table <- anova(rda_t1,by="margin")
#rda_t1_table <- anova.cca(rda_t1, by = "margin", permutations=999)
#kable(rda_t1_table, digits = 3) 

```

#relative abundance bar plot(family level)

#select top 15 families for all data
```{r}

taxonomy_all<-bind_rows(taxonomy_t0[,1:9],taxonomy_t1[,1:9])%>% distinct(ASV_ID, .keep_all = TRUE)
sample_meta_data_all<-rbind(sample_meta_data_t0,sample_meta_data_t1)

p_all<- rbind(asvs_t0,asvs_t1) %>%
  inner_join(taxonomy_all, by = 'ASV_ID') %>%
  inner_join(sample_meta_data_all, by = 'sample')%>%
  group_by(Family, sample) %>%
  # Calculate the relative abundance of each family in each sample
  summarise(relab = sum(relab)) %>%
  # Calculate the *mean* relative abundance of each family over the samples
  summarise(mean_relab = sum(relab)) %>%
  ungroup() %>%
  # Filter out non-assigned family
  filter(!is.na(Family))%>%
  # Select the top 15
  top_n(16, mean_relab)


p_all

p_all_without_unidentified<-p_all[-16,]

```



#t0
```{r}


#depth
depth_Family_t0<-asvs_t0[,-4]%>% 
  filter(count > 0)%>%
  # Add metadata
 inner_join(sample_meta_data_t0, by = 'sample')%>%
  inner_join(taxonomy_t0, by = 'ASV_ID')%>%
  group_by(Bay, Depth)%>%
  # Calculate the relative abundance of each family in each site
  mutate(relab = count/sum(count))%>%
  ungroup()%>% 
  left_join(
    p_all %>% 
      transmute(Family, top_15_Family = Family),
    by = 'Family'
  ) %>%
  replace_na(list('top_15_Family' = 'Other_Families'))

ra_plot_table_t0<-depth_Family_t0%>%
  group_by(Bay, Depth,top_15_Family)%>%
  summarise(relab=sum(relab))

# Get all family names except Other_Families and unidentified
family_levels <- ra_plot_table_t0 %>%
  filter(!top_15_Family %in% c("Other_Families", "unidentified")) %>%
  pull(top_15_Family) %>%
  unique() %>%
  sort()

# Append Other_Families and unidentified at the end
family_levels <- c(family_levels, "Other_Families", "unidentified")

# Apply factor ordering to the plot table
ra_plot_table_t0$top_15_Family <- factor(ra_plot_table_t0$top_15_Family, 
                                          levels = family_levels)

named_colors <- setNames(col_vector[1:length(family_levels)], family_levels)

# Get the colour currently assigned to unidentified by position
unidentified_color <- named_colors["unidentified"]

# Assign that colour to Other_Families, then set unidentified to light grey
named_colors["Other_Families"] <- unidentified_color
named_colors["unidentified"] <- "lightgrey"
ra_plot_table_t0$Bay <- factor(ra_plot_table_t0$Bay, 
                                           levels = c("heated", "control"))

ra_plot_table_t0_only_top15<-ra_plot_table_t0[ra_plot_table_t0$top_15_Family != 'Other_Families',  ]
ra_plot_table_t0_only_top15<-ra_plot_table_t0_only_top15[ra_plot_table_t0_only_top15$top_15_Family !=  'unidentified', ]
ra_plot_table_t0_only_top15$Bay <- factor(ra_plot_table_t0_only_top15$Bay, 
                                           levels = c("heated", "control"))

ra_plot_t0_all<-ggplot(ra_plot_table_t0,aes(x=Depth, y=relab, fill=top_15_Family)) +
  facet_wrap(~Bay, ncol= 2, 
           labeller = labeller(Bay = c("heated" = "T0 Heated bay", 
                                       "control" = "T0 Control bay")))+
  geom_col() +
  theme_bw() +
  scale_fill_manual(values = named_colors)+
  scale_x_discrete(limits=rev)+
  coord_flip()+
  labs(x="Sediment depth (cm)", y=element_blank()) +
  theme_consistent2+
  theme(legend.position="bottom")

ra_plot_t0_all


ra_plot_t0_seleted<-ggplot(ra_plot_table_t0_only_top15,aes(x=Depth, y=relab, fill=top_15_Family)) +
  facet_wrap(~Bay, ncol= 2, 
           labeller = labeller(Bay = c("heated" = "T0 Heated bay", 
                                       "control" = "T0 Control bay")))+
  geom_col() +
  theme_bw() +
  scale_fill_manual(values = named_colors)+
  ylim(0,0.4)+
  scale_x_discrete(limits=rev)+
  coord_flip()+
  labs(x="Sediment depth (cm)", y=element_blank()) +
  theme_consistent2+
  theme(legend.position="bottom")

ra_plot_t0_seleted

```

#t1

#t1 heated bay
```{r}


#depth
depth_Family_t1<-asvs_t1[,-4]%>% 
  filter(count > 0)%>%
  # Add metadata
 inner_join(sample_meta_data_t1, by = 'sample')%>%
  inner_join(taxonomy_t1, by = 'ASV_ID')%>%
  group_by(Bay, Core, Depth)%>%
  # Calculate the relative abundance of each family in each site
  mutate(relab = count/sum(count))%>%
  ungroup()%>% 
  left_join(
    p_all %>% 
      transmute(Family, top_15_Family = Family),
    by = 'Family'
  ) %>%
  replace_na(list('top_15_Family' = 'Other_Families'))

ra_plot_table_t1_heated<-depth_Family_t1%>%
  filter(Bay == "heated")%>%
  group_by(Core, Depth,top_15_Family)%>%
  summarise(relab=sum(relab))

# Get all family names except Other_Families and unidentified
family_levels <- ra_plot_table_t1_heated %>%
  filter(!top_15_Family %in% c("Other_Families", "unidentified")) %>%
  pull(top_15_Family) %>%
  unique() %>%
  sort()

# Append Other_Families and unidentified at the end
family_levels <- c(family_levels, "Other_Families", "unidentified")

# Apply factor ordering to the plot table
ra_plot_table_t1_heated$top_15_Family <- factor(ra_plot_table_t1_heated$top_15_Family, 
                                          levels = family_levels)

  
ra_plot_t1_heated_all<-ggplot(ra_plot_table_t1_heated,aes(x=Depth, y=relab, fill=top_15_Family)) +
  facet_wrap(~Core, ncol= 2, 
           labeller = labeller(Core = c("Resident" = "T1 Heated Resident", 
                                       "Translocated" = "T1 Translocated C->H")))+
  geom_col() +
  scale_fill_manual(values = named_colors)+
  scale_x_discrete(limits=rev)+
  coord_flip()+
  labs(x="Sediment depth (cm)", y=element_blank())+
  theme_consistent2+
  theme(legend.position="bottom")

ra_plot_t1_heated_all



ra_plot_table_t1_heated_only_top15<-ra_plot_table_t1_heated[ra_plot_table_t1_heated$top_15_Family != 'Other_Families',  ]
ra_plot_table_t1_heated_only_top15<-ra_plot_table_t1_heated_only_top15[ra_plot_table_t1_heated_only_top15$top_15_Family !=  'unidentified', ]


ra_plot_t1_heated_seleted<-ggplot(ra_plot_table_t1_heated_only_top15,aes(x=Depth, y=relab, fill=top_15_Family)) +
  facet_wrap(~Core, ncol= 2, 
           labeller = labeller(Core = c("Resident" = "T1 Heated Resident", 
                                       "Translocated" = "T1 Translocated C->H")))+
  geom_col() +
  scale_fill_manual(values = named_colors)+
  scale_x_discrete(limits=rev)+
  coord_flip()+
  ylim(0,0.4)+
  labs(x="Sediment depth (cm)", y=element_blank())+
  theme_consistent2+
  theme(legend.position="bottom")

ra_plot_t1_heated_seleted


```

#t1 control bay
```{r}

#depth
depth_Family_t1<-asvs_t1[,-4]%>% 
  filter(count > 0)%>%
  # Add metadata
 inner_join(sample_meta_data_t1, by = 'sample')%>%
  inner_join(taxonomy_t1, by = 'ASV_ID')%>%
  group_by(Bay, Core, Depth)%>%
  # Calculate the relative abundance of each family in each site
  mutate(relab = count/sum(count))%>%
  ungroup()%>% 
  left_join(
    p_all%>% 
      transmute(Family, top_15_Family = Family),
    by = 'Family'
  ) %>%
  replace_na(list('top_15_Family' = 'Other_Families'))

ra_plot_table_t1_control<-depth_Family_t1%>%
  filter(Bay == "control")%>%
  group_by(Core, Depth,top_15_Family)%>%
  summarise(relab=sum(relab))

# Get all family names except Other_Families and unidentified
family_levels <- ra_plot_table_t1_control %>%
  filter(!top_15_Family %in% c("Other_Families", "unidentified")) %>%
  pull(top_15_Family) %>%
  unique() %>%
  sort()

# Append Other_Families and unidentified at the end
family_levels <- c(family_levels, "Other_Families", "unidentified")

# Apply factor ordering to the plot table
ra_plot_table_t1_control$top_15_Family <- factor(ra_plot_table_t1_control$top_15_Family, 
                                          levels = family_levels)

  
ra_plot_t1_control_all<-ggplot(ra_plot_table_t1_control,aes(x=Depth, y=relab, fill=top_15_Family)) +
  facet_wrap(~Core, ncol= 2, 
           labeller = labeller(Core = c("Resident" = "T1 Control Resident", 
                                       "Translocated" = "T1 Translocated H->C")))+
  geom_col() +
  scale_fill_manual(values = named_colors)+
  scale_x_discrete(limits=rev)+
  coord_flip()+
  labs(x="Sediment depth (cm)", y=element_blank())+
  theme_consistent2+
  theme(legend.position="bottom")

ra_plot_t1_control_all


ra_plot_table_t1_control_only_top15<-ra_plot_table_t1_control[ra_plot_table_t1_control$top_15_Family != 'Other_Families',  ]
ra_plot_table_t1_control_only_top15<-ra_plot_table_t1_control_only_top15[ra_plot_table_t1_control_only_top15$top_15_Family !=  'unidentified', ]


ra_plot_t1_control_seleted<-ggplot(ra_plot_table_t1_control_only_top15,aes(x=Depth, y=relab, fill=top_15_Family)) +
  facet_wrap(~Core, ncol= 2, 
           labeller = labeller(Core = c("Resident" = "T1 Control Resident", 
                                       "Translocated" = "T1 Translocated H->C")))+
  geom_col() +
  scale_fill_manual(values = named_colors)+
  scale_x_discrete(limits=rev)+
  coord_flip()+
  ylim(0,0.4)+
  labs(x="Sediment depth (cm)", y=element_blank())+
  theme_consistent2+
  theme(legend.position="bottom")

ra_plot_t1_control_seleted

```


#try mixed barplots
```{r}

top15_mix<-ggarrange(
  ra_plot_t0_seleted,
  ra_plot_t1_heated_seleted, 
  ra_plot_t1_control_seleted,
  labels = c("a", "b","c"), 
  font.label = list(size = 18),
  nrow  = 3,
  common.legend = TRUE,
  legend = "bottom"
)

top15_mix

ggsave("top15_mix.pdf", plot = top15_mix,width = 25, height =30,units = "cm")

supp_all_mix<-ggarrange(
  ra_plot_t0_all,
  ra_plot_t1_heated_all, 
  ra_plot_t1_control_all,
  labels = c("a", "b","c"), 
  font.label = list(size = 18),
  nrow  = 3,
  common.legend = TRUE,
  legend = "bottom"
)

ggsave("supp_all_mix.pdf", plot = supp_all_mix,width = 25, height =30,units = "cm")

supp_all_mix

```


## gllvm
## family abundance analysis
```{r gllvm T0} 
# Load libraries
library(dplyr)
library(tidyr)
library(readr)
library(phyloseq)
library(gllvm)
library(ggplot2)
library(tibble)
library(ggh4x)

# Prepare phyloseq object
ps_T0 <- phyloseq(
  otu_table(otu_mat_T0, taxa_are_rows = TRUE),
  tax_table(tax_mat_T0),
  sample_data(samples_df_T0)
)

# Aggregate to family level
ps_T0_family <- tax_glom(ps_T0, taxrank = "Family")

# Select top 15 families
top_15_families <- c(
  "34-128", "Anaerolineaceae", "Caldilineaceae", "Chromatiaceae", "Cyanobiaceae",
  "Desulfocapsaceae", "Flavobacteriaceae", "Ilumatobacteraceae", "JAAYZQ01",
  "Pirellulaceae", "Thiobacillaceae", "UBA11574", "UBA4823", "VadinHA17"
)

otu_count_family_T0 <- as.data.frame(otu_table(ps_T0_family)) %>%
  mutate(Family = as.vector(tax_table(ps_T0_family)[, "Family"]))

otu_count_family_T0_15 <- otu_count_family_T0 %>%
  mutate(Family = ifelse(Family %in% top_15_families, Family, "Other")) %>%
  group_by(Family) %>%
  summarise(across(where(is.numeric), sum), .groups = "drop")

otu_mat_T0_15 <- otu_count_family_T0_15 %>%
  column_to_rownames("Family") %>%
  t() %>%
  as.matrix()

# Prepare metadata
sample_meta_T0 <- sample_data(ps_T0_family) %>%
  data.frame() %>%
  mutate(
    Core_ID = factor(Core_ID),
    Bay = factor(Bay),
    Core = factor(Core),
    Depth = as.numeric(as.character(Depth))
  ) %>%
  filter(rownames(.) %in% rownames(otu_mat_T0_15))

# Validate matching
stopifnot(all(rownames(otu_mat_T0_15) == rownames(sample_meta_T0)))

# Define covariates and design
X_T0 <- sample_meta_T0[, c("Bay", "Depth")] %>%
  mutate(
    Bay = factor(Bay, levels = c("control", "heated")),
    Depth = as.numeric(scale(Depth))  # standardise
  )

study_design_T0 <- data.frame(Core_ID = sample_meta_T0$Core_ID)

# Define offset
offset_vec_T0 <- log(rowSums(otu_mat_T0_15))

# Fit gllvm model
fit_T0_top15 <- gllvm(
  y = otu_mat_T0_15,
  X = X_T0,
  formula = ~ Bay * Depth,
  family = "negative.binomial",
  offset = offset_vec_T0,
  studyDesign = study_design_T0,
  row.eff = ~(1 | Core_ID),
  num.lv = 0
)
fit_T0_top15_1 <- gllvm(
  y = otu_mat_T0_15,
  X = X_T0,
  formula = ~ Bay * Depth,
  family = "negative.binomial",
  offset = offset_vec_T0,
  studyDesign = study_design_T0,
  row.eff = ~(1 | Core_ID),
  num.lv = 1
)
fit_T0_top15_2 <- gllvm(
  y = otu_mat_T0_15,
  X = X_T0,
  formula = ~ Bay * Depth,
  family = "negative.binomial",
  offset = offset_vec_T0,
  studyDesign = study_design_T0,
  row.eff = ~(1 | Core_ID),
  num.lv = 2
)

#use AIC to select number of latent variables
AIC(fit_T0_top15)
AIC(fit_T0_top15_1)# 1 or 2 are best fit
AIC(fit_T0_top15_2)

#residual analysis
par(mfrow = c(1,2))
plot(fit_T0_top15_1, which = 1:2)

#visualize species correlations
cr_T0 <- getResidualCor(fit_T0_top15_1)
library(corrplot); library(gclus)
corrplot(cr_T0[order.single(cr_T0), order.single(cr_T0)], diag = FALSE, type = "lower", 
         method = "square", tl.cex = 1, tl.srt = 45, tl.col = "red")
         
# Summarise and export
summary_gllvm_T0 <- summary(fit_T0_top15_1)
gllvm_results_T0 <- as.data.frame(summary_gllvm_T0$Coef.tableX)
write.csv(gllvm_results_T0, "gllvm_results_T0_top15.csv", row.names = TRUE)

# Plot latent space
gllvm::ordiplot(fit_T0_top15_1)
coefplot(fit_T0_top15_1)

# Plot effects per family
coefs_T0 <- coef(fit_T0_top15_1)$Xcoef
ses_T0 <- fit_T0_top15_1$sd$Xcoef
beta0_T0 <- coef(fit_T0_top15_1)$Intercept
taxa_names_T0 <- intersect(rownames(coefs_T0), rownames(ses_T0))

bay_levels_T0 <- levels(X_T0$Bay)
depth_seq_T0 <- seq(min(X_T0$Depth), max(X_T0$Depth), length.out = 100)

newdata_T0 <- expand.grid(Bay = bay_levels_T0, Depth = depth_seq_T0) %>%
  mutate(
    Bay = factor(Bay, levels = bay_levels_T0),
    Depth = as.numeric(Depth)
  )

Xmat_no_int_T0 <- model.matrix(~ Bay * Depth, data = newdata_T0)[, -1, drop = FALSE]

pred_list_T0 <- lapply(taxa_names_T0, function(fam) {
  beta <- as.numeric(coefs_T0[fam, ])
  beta_se <- as.numeric(ses_T0[fam, ])
  intercept <- beta0_T0[fam]
  
  if (any(is.na(beta)) || any(is.na(beta_se))) return(NULL)
  
  eta <- intercept + Xmat_no_int_T0 %*% beta
  se_eta <- sqrt(rowSums(sweep(Xmat_no_int_T0, 2, beta_se, `*`)^2))
  
  tibble(
    Bay = newdata_T0$Bay,
    Depth = newdata_T0$Depth,
    Family = fam,
    predicted = exp(eta),
    lower = exp(eta - 1.96 * se_eta),
    upper = exp(eta + 1.96 * se_eta)
  )
})

pred_df_T0 <- bind_rows(pred_list_T0)


# Order families with Other last
family_order <- sort(unique(pred_df_T0$Family))
family_order <- c(family_order[family_order != "Other"], "Other")
pred_df_T0$Family <- factor(pred_df_T0$Family, levels = family_order)

# Back-transform standardised depth to actual depth for x axis
depth_actual <- seq(min(sample_meta_T0$Depth), max(sample_meta_T0$Depth), length.out = 100)
depth_scaled <- seq(min(X_T0$Depth), max(X_T0$Depth), length.out = 100)

# Add actual depth to pred_df_T0
pred_df_T0$Depth_actual <- depth_actual[match(round(pred_df_T0$Depth, 6), 
                                               round(depth_scaled, 6))]
# Safer back-transform using scale attributes
depth_mean <- attr(scale(sample_meta_T0$Depth), "scaled:center")
depth_sd   <- attr(scale(sample_meta_T0$Depth), "scaled:scale")
pred_df_T0$Depth_actual <- pred_df_T0$Depth * depth_sd + depth_mean

# Plot
p_gllvm_T0_15_cm <- ggplot(pred_df_T0, aes(x = Depth_actual, y = predicted, 
                                          colour = Bay, fill = Bay)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_wrap(~ Family, scales = "free_y", ncol = 4) +
  scale_colour_manual(values = c("control" = "#2166AC", "heated" = "#D6191B")) +
  scale_fill_manual(values = c("control" = "#2166AC", "heated" = "#D6191B")) +
  labs(x = "Depth (cm)", y = "Predicted abundance", colour = "Bay", fill = "Bay") +
  theme_bw() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 8),
    plot.title = element_text(size = 12),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    strip.text = element_text(size = 7),
    panel.border = element_rect(linewidth = 1.25),
    strip.background = element_rect(fill = "white", colour = NA,linewidth = 0.5),
    legend.position = c(0.88, 0.12),  # bottom right of plotting area
    legend.background = element_rect(fill = "white", colour = NA, linewidth = 0.5),
    legend.margin = margin(4, 6, 4, 6)
  )

p_gllvm_T0_15_cm
ggsave("p_gllvm_T0_top15_cm.jpg", plot = p_gllvm_T0_15_cm, width = 17, height = 12, units = "cm", dpi = 1200)


# Plot
p_gllvm_T0_15 <- ggplot(pred_df_T0, aes(x = Depth, y = predicted, colour = Bay, fill = Bay)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_wrap(~ Family, scales = "free_y") +
  scale_colour_manual(values = c("control" = "#2166AC", "heated" = "#D6191B")) +
  scale_fill_manual(values = c("control" = "#2166AC", "heated" = "#D6191B")) +
  labs(x = "Depth", y = "Predicted abundance", colour = "Bay", fill = "Bay") +
  theme_bw() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 8),
    plot.title = element_text(size = 12),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    strip.text = element_text(size = 7),
    panel.border = element_rect(linewidth = 1.25),
    strip.background = element_rect(fill = "white", colour = NA,linewidth = 0.5),
    legend.position = c(0.88, 0.12),  # bottom right of plotting area
    legend.background = element_rect(fill = "white", colour = NA, linewidth = 0.5),
    legend.margin = margin(4, 6, 4, 6)
  )

p_gllvm_T0_15

ggsave("p_gllvm_T0_top15.jpg", plot = p_gllvm_T0_15, width = 17, height = 12, units = "cm", dpi = 1200)


## coefficient plot in ggplot to modify
library(tidyverse)
# Reshape coefficients to long format
coef_T0_long <- as.data.frame(as.table(coefs_T0)) %>%
  rename(Family = Var1, Variable = Var2, Estimate = Freq) %>%
  mutate(
    SE = as.vector(ses_T0)
  )

# Calculate 95% CI and significance
coef_T0_long <- coef_T0_long %>%
  mutate(
    lower = Estimate - 1.96 * SE,
    upper = Estimate + 1.96 * SE,
    sig = ifelse(lower * upper > 0, "yes", "no")  # significant if CI does not cross 0
  )

# Clean variable names
coef_T0_long$Variable <- coef_T0_long$Variable %>%
  str_replace("Bayheated:Depth", "Bay × Depth") %>%
  str_replace("Bayheated", "Bay") %>%
  str_replace("Depth", "Depth")
# Set Variable as a factor with desired order
coef_T0_long$Variable <- factor(coef_T0_long$Variable, levels = c("Bay", "Depth", "Bay × Depth"))

# Reorder families 
family_order_T0 <- c(
  sort(unique(as.character(coef_T0_long$Family[coef_T0_long$Family != "Other"]))),
  "Other"
)
coef_T0_long$Family <- factor(coef_T0_long$Family, levels = rev(family_order_T0))


p_coef_T0 <- ggplot(coef_T0_long, aes(x = Estimate, y = Family, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0.2,
                 position = position_dodge(width = 0.5)) +
  geom_point(shape = 4, size = 2, position = position_dodge(width = 0.5)) +
  facet_wrap(~ Variable,  ncol = 3) +
  scale_colour_manual(values = c("yes" = "black", "no" = "grey70")) +
  labs(x = "Estimate", y = NULL, colour = "Significance") +
  theme_bw() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 9),
    plot.title = element_text(size = 12),
    strip.text = element_text(size = 11),
    panel.border = element_rect(linewidth = 1.25),
    strip.background = element_rect(fill = "white", colour = NA,linewidth = 0.5),
    legend.position = "none",
    panel.spacing = unit(1, "lines"))

p_coef_T0

# Save
ggsave("p_coef_T0_by_variable.pdf", plot = p_coef_T0, width = 20, height = 16, units = "cm", dpi = 1200)
```

```{r gllvm T1} 
# Make phyloseq object
ps_T1 <- phyloseq(
  otu_table(otu_mat_T1, taxa_are_rows = TRUE),
  tax_table(tax_mat_T1),
  sample_data(samples_df_T1)
)

# Aggregate to family level
ps_T1_family <- tax_glom(ps_T1, taxrank = "Family")

# Select top 15 families
top_15_families <- c(
  "34-128", "Anaerolineaceae", "Caldilineaceae", "Chromatiaceae", "Cyanobiaceae",
  "Desulfocapsaceae", "Flavobacteriaceae", "Ilumatobacteraceae", "JAAYZQ01",
  "Pirellulaceae", "Thiobacillaceae", "UBA11574", "UBA4823", "VadinHA17"
)

otu_count_family_T1 <- as.data.frame(otu_table(ps_T1_family)) %>%
  mutate(Family = as.vector(tax_table(ps_T1_family)[, "Family"]))

otu_count_family_T1_15 <- otu_count_family_T1 %>%
  mutate(Family = ifelse(Family %in% top_15_families, Family, "Other")) %>%
  group_by(Family) %>%
  summarise(across(where(is.numeric), sum), .groups = "drop")

otu_mat_T1_15 <- otu_count_family_T1_15 %>%
  column_to_rownames("Family") %>%
  t() %>%
  as.matrix()

# Prepare metadata
sample_meta_T1 <- sample_data(ps_T1_family) %>%
  data.frame() %>%
  mutate(
    Core_ID = factor(Core_ID),
    Bay = factor(Bay),
    Core = factor(Core),
    Depth = as.numeric(as.character(Depth))
  ) %>%
  filter(rownames(.) %in% rownames(otu_mat_T1_15))

# Validate matching
stopifnot(all(rownames(otu_mat_T1_15) == rownames(sample_meta_T1)))

# Define covariates and design
X_T1 <- sample_meta_T1[, c("Bay", "Core", "Depth")] %>%
  mutate(
    Bay = factor(Bay),
    Core = factor(Core),
    Depth = as.numeric(scale(Depth))  # standardise
  )

study_design_T1 <- data.frame(Core_ID = sample_meta_T1$Core_ID)

# Define offset
offset_vec_T1 <- log(rowSums(otu_mat_T1_15))

# Fit gllvm model
fit_T1_top15 <- gllvm(
  y = otu_mat_T1_15,
  X = X_T1,
  formula = ~ Bay * Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1,
  studyDesign = study_design_T1,
  row.eff = ~(1 | Core_ID),
  num.lv = 0
)
fit_T1_top15_1 <- gllvm(
  y = otu_mat_T1_15,
  X = X_T1,
  formula = ~ Bay * Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1,
  studyDesign = study_design_T1,
  row.eff = ~(1 | Core_ID),
  num.lv = 1
)
fit_T1_top15_2 <- gllvm(
  y = otu_mat_T1_15,
  X = X_T1,
  formula = ~ Bay * Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1,
  studyDesign = study_design_T1,
  row.eff = ~(1 | Core_ID),
  num.lv = 2
)

#use AIC to select number of latent variables
AIC(fit_T1_top15)
AIC(fit_T1_top15_1)# 1 or 2 are best fit
AIC(fit_T1_top15_2)

#residual analysis
par(mfrow = c(1,2))
plot(fit_T1_top15_1, which = 1:2)

#visualize species correlations
cr_T1 <- getResidualCor(fit_T1_top15_1)
library(corrplot); library(gclus)
corrplot(cr_T1[order.single(cr_T1), order.single(cr_T1)], diag = FALSE, type = "lower", 
         method = "square", tl.cex = 1, tl.srt = 45, tl.col = "red")
         
# Summarise and export
summary_gllvm_T1 <- summary(fit_T1_top15_1)
gllvm_results_T1 <- as.data.frame(summary_gllvm_T1$Coef.tableX)
write.csv(gllvm_results_T1, "gllvm_results_T1_top15.csv", row.names = TRUE)

# Plot latent space
gllvm::ordiplot(fit_T1_top15_1)

#coefficient plot!!
coefplot(fit_T1_top15_1, mfrow=c(2,4))

# Plot effects per family
coefs_T1 <- coef(fit_T1_top15_1)$Xcoef
ses_T1 <- fit_T1_top15_1$sd$Xcoef
beta0_T1 <- coef(fit_T1_top15_1)$Intercept
taxa_names_T1 <- intersect(rownames(coefs_T1), rownames(ses_T1))

bay_levels_T1 <- levels(X_T1$Bay)
core_levels_T1<- levels(X_T1$Core)
depth_seq_T1 <- seq(min(X_T1$Depth), max(X_T1$Depth), length.out = 100)

newdata_T1 <- expand.grid(Bay = bay_levels_T1, Core = core_levels_T1, Depth = depth_seq_T1) %>%
  mutate(
    Bay = factor(Bay, levels = bay_levels_T1),
    Core = factor(Core, levels = core_levels_T1),
    Depth = as.numeric(Depth)
  )

Xmat_no_int_T1 <- model.matrix(~ Bay * Core * Depth, data = newdata_T1)[, -1, drop = FALSE]

pred_list_T1 <- lapply(taxa_names_T1, function(fam) {
  beta <- as.numeric(coefs_T1[fam, ])
  beta_se <- as.numeric(ses_T1[fam, ])
  intercept <- beta0_T1[fam]
  
  if (any(is.na(beta)) || any(is.na(beta_se))) return(NULL)
  
  eta <- intercept + Xmat_no_int_T1 %*% beta
  se_eta <- sqrt(rowSums(sweep(Xmat_no_int_T1, 2, beta_se, `*`)^2))
  
  tibble(
    Bay = newdata_T1$Bay,
    Core = newdata_T1$Core,
    Depth = newdata_T1$Depth,
    Family = fam,
    predicted = exp(eta),
    lower = exp(eta - 1.96 * se_eta),
    upper = exp(eta + 1.96 * se_eta)
  )
})

pred_df_T1 <- bind_rows(pred_list_T1)

# Plot

# Order families with Other last
family_order <- sort(unique(pred_df_T1$Family))
family_order <- c(family_order[family_order != "Other"], "Other")
pred_df_T1$Family <- factor(pred_df_T1$Family, levels = family_order)

# Back-transform standardised depth to actual depth for x axis
depth_actual <- seq(min(sample_meta_T1$Depth), max(sample_meta_T1$Depth), length.out = 100)
depth_scaled <- seq(min(X_T1$Depth), max(X_T1$Depth), length.out = 100)

# Add actual depth to pred_df_T1
pred_df_T1$Depth_actual <- depth_actual[match(round(pred_df_T1$Depth, 6), 
                                               round(depth_scaled, 6))]
# Safer back-transform using scale attributes
depth_mean <- attr(scale(sample_meta_T1$Depth), "scaled:center")
depth_sd   <- attr(scale(sample_meta_T1$Depth), "scaled:scale")
pred_df_T1$Depth_actual <- pred_df_T1$Depth * depth_sd + depth_mean


p_gllvm_T1_15 <- ggplot(pred_df_T1, aes(x = Depth_actual, y = predicted, colour = Core, fill = Core)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_nested(Family ~ Bay, scales = "free_y", space = "free_x") +  
  scale_colour_manual(values = c("red4", "cyan4")) +
  scale_fill_manual(values = c("red4", "cyan4")) +
  labs(x = "Depth", y = "Predicted abundance") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 10),
    legend.position = "none"
  )

p_gllvm_T1_15

p_gllvm_T1_15_control <- ggplot(pred_df_T1[pred_df_T1$Bay=="control",], aes(x = Depth_actual, y = predicted, colour = Core, fill = Core)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_wrap(~ Family, scales = "free_y") +
  scale_colour_manual(values = c("#2166AC", "#F57C00")) +
  scale_fill_manual(values = c("#2166AC", "#F57C00")) +
  labs(x = "Depth (cm)", y = "Predicted abundance") +
  theme_bw() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 8),
    plot.title = element_text(size = 12),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    strip.text = element_text(size = 7),
    panel.border = element_rect(linewidth = 1.25),
    strip.background = element_rect(fill = "white", colour = NA,linewidth = 0.5),
    legend.position = c(0.88, 0.12),  # bottom right of plotting area
    legend.background = element_rect(fill = "white", colour = NA, linewidth = 0.5),
    legend.margin = margin(4, 6, 4, 6))

p_gllvm_T1_15_control
ggsave("p_gllvm_T1_control_top15.pdf", plot = p_gllvm_T1_15_control, width = 17, height = 12, units = "cm", dpi = 1200)

p_gllvm_T1_15_heated <- ggplot(pred_df_T1[pred_df_T1$Bay=="heated",], aes(x = Depth_actual, y = predicted, colour = Core, fill = Core)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_wrap(~ Family, scales = "free_y") +
  scale_colour_manual(values = c("#D6191B", "#00ACC1")) +
  scale_fill_manual(values = c("#D6191B", "#00ACC1")) +
  labs(x = "Depth (cm)", y = "Predicted abundance") +
  theme_bw() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 8),
    plot.title = element_text(size = 12),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    strip.text = element_text(size = 7),
    panel.border = element_rect(linewidth = 1.25),
    strip.background = element_rect(fill = "white", colour = NA,linewidth = 0.5),
    legend.position = c(0.88, 0.12),  # bottom right of plotting area
    legend.background = element_rect(fill = "white", colour = NA, linewidth = 0.5),
    legend.margin = margin(4, 6, 4, 6))

p_gllvm_T1_15_heated
ggsave("p_gllvm_T1_heated_top15.pdf", plot = p_gllvm_T1_15_heated, width = 17, height = 12, units = "cm", dpi = 1200)


```

```{r gllvm T1 bays separate} 
# Subset metadata and OTU table by Bay
sample_meta_T1_control <- sample_meta_T1 %>% filter(Bay == "control")
sample_meta_T1_heated <- sample_meta_T1 %>% filter(Bay == "heated")

otu_mat_T1_15_control <- otu_mat_T1_15[rownames(sample_meta_T1_control), ]
otu_mat_T1_15_heated <- otu_mat_T1_15[rownames(sample_meta_T1_heated), ]

# Prepare design matrices for each Bay
X_T1_control <- sample_meta_T1_control %>%
  mutate(
    Core = factor(Core),
    Depth = as.numeric(scale(Depth)) # standardise within Bay
  )

X_T1_heated <- sample_meta_T1_heated %>%
  mutate(
    Core = factor(Core),
    Depth = as.numeric(scale(Depth))
  )

# Study design
study_design_T1_control <- data.frame(Core_ID = sample_meta_T1_control$Core_ID)
study_design_T1_heated <- data.frame(Core_ID = sample_meta_T1_heated$Core_ID)

# Define offsets
offset_vec_T1_control <- log(rowSums(otu_mat_T1_15_control))
offset_vec_T1_heated <- log(rowSums(otu_mat_T1_15_heated))

# Fit models separately
fit_T1_control <- gllvm(
  y = otu_mat_T1_15_control,
  X = X_T1_control,
  formula = ~ Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1_control,
  studyDesign = study_design_T1_control,
  row.eff = ~(1 | Core_ID),
  num.lv = 1
)

fit_T1_heated <- gllvm(
  y = otu_mat_T1_15_heated,
  X = X_T1_heated,
  formula = ~ Core * Depth,
  family = "negative.binomial",
  offset = offset_vec_T1_heated,
  studyDesign = study_design_T1_heated,
  row.eff = ~(1 | Core_ID),
  num.lv = 1
)

# Compare model AICs 
AIC(fit_T1_control)
AIC(fit_T1_heated)

# Summarise and export
summary_control <- summary(fit_T1_control)
summary_heated <- summary(fit_T1_heated)

write.csv(as.data.frame(summary_control$Coef.tableX), "gllvm_results_T1_control.csv", row.names = TRUE)
write.csv(as.data.frame(summary_heated$Coef.tableX), "gllvm_results_T1_heated.csv", row.names = TRUE)

# check plots of residuals
par(mfrow = c(1, 2))
plot(fit_T1_control, which = 1:2)
plot(fit_T1_heated, which = 1:2)

# latent space plots
gllvm::ordiplot(fit_T1_control, main = "Control Bay")
gllvm::ordiplot(fit_T1_heated, main = "Heated Bay")

#coefficient plots!!
coefplot(fit_T1_control)
coefplot(fit_T1_heated)

#prediction plots
# Define prediction grid for each Bay
core_levels_control <- levels(X_T1_control$Core)
core_levels_heated <- levels(X_T1_heated$Core)
depth_seq_control <- seq(min(X_T1_control$Depth), max(X_T1_control$Depth), length.out = 100)
depth_seq_heated <- seq(min(X_T1_heated$Depth), max(X_T1_heated$Depth), length.out = 100)
newdata_control <- expand.grid(Core = core_levels_control, Depth = depth_seq_control)
newdata_heated <- expand.grid(Core = core_levels_heated, Depth = depth_seq_heated)
Xmat_control <- model.matrix(~ Core * Depth, data = newdata_control)[, -1, drop = FALSE]
Xmat_heated <- model.matrix(~ Core * Depth, data = newdata_heated)[, -1, drop = FALSE]
# Extract coefficients
coefs_control <- coef(fit_T1_control)$Xcoef
ses_control <- fit_T1_control$sd$Xcoef
beta0_control <- coef(fit_T1_control)$Intercept
coefs_heated <- coef(fit_T1_heated)$Xcoef
ses_heated <- fit_T1_heated$sd$Xcoef
beta0_heated <- coef(fit_T1_heated)$Intercept
taxa_names_control <- intersect(rownames(coefs_control), rownames(ses_control))
taxa_names_heated <- intersect(rownames(coefs_heated), rownames(ses_heated))
# Predict function
predict_gllvm <- function(coefs, ses, intercepts, taxa_names, Xmat, newdata, Bay_label) {
  lapply(taxa_names, function(fam) {
    beta <- as.numeric(coefs[fam, ])
    beta_se <- as.numeric(ses[fam, ])
    intercept <- intercepts[fam]
    if (any(is.na(beta)) || any(is.na(beta_se))) return(NULL)
    eta <- intercept + Xmat %*% beta
    se_eta <- sqrt(rowSums(sweep(Xmat, 2, beta_se, `*`)^2))
    tibble(
      Core = newdata$Core,
      Depth = newdata$Depth,
      Family = fam,
      Bay = Bay_label,
      predicted = exp(eta),
      lower = exp(eta - 1.96 * se_eta),
      upper = exp(eta + 1.96 * se_eta)
    )
  }) %>% bind_rows()
}
# Predict
pred_df_control <- predict_gllvm(coefs_control, ses_control, beta0_control, taxa_names_control, Xmat_control, newdata_control, Bay_label = "control")
pred_df_heated <- predict_gllvm(coefs_heated, ses_heated, beta0_heated, taxa_names_heated, Xmat_heated, newdata_heated, Bay_label = "heated")

p_gllvm_T1_heated <- ggplot(pred_df_heated, aes(x = Depth, y = predicted, colour = Core, fill = Core)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_nested(Family ~ Bay, scales = "free_y", space = "free_x") +  
  scale_colour_manual(values = c("#D6191B", "#00ACC1")) +
  scale_fill_manual(values = c("#D6191B", "#00ACC1")) +
  labs(x = "Depth", y = "Predicted abundance") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 10),
    legend.position = "none"
  )

p_gllvm_T1_heated


p_gllvm_T1_control <- ggplot(pred_df_control, aes(x = Depth, y = predicted, colour = Core, fill = Core)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, colour = NA) +
  facet_nested(Family ~ Bay, scales = "free_y", space = "free_x") +  
  scale_colour_manual(values = c("#2166AC", "#F57C00")) +
  scale_fill_manual(values = c("#2166AC", "#F57C00")) +
  labs(x = "Depth", y = "Predicted abundance") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 10),
    legend.position = "none"
  )

p_gllvm_T1_control

#Combined Coefficient Plot
library(tidyverse)
coef_control_long <- as.data.frame(as.table(coefs_control)) %>%
  rename(Family = Var1, Variable = Var2, Estimate = Freq) %>%
  mutate(
    SE = as.vector(ses_control),
    Bay = "control"
  )
coef_heated_long <- as.data.frame(as.table(coefs_heated)) %>%
  rename(Family = Var1, Variable = Var2, Estimate = Freq) %>%
  mutate(
    SE = as.vector(ses_heated),
    Bay = "heated"
  )
# Combine
coef_long <- bind_rows(coef_control_long, coef_heated_long)
# Keep only predictors of interest
coef_long <- coef_long %>%
  filter(str_detect(Variable, "Core|Depth"))
# Calculate 95% CIs and significance
coef_long <- coef_long %>%
  mutate(
    lower = Estimate - 1.96 * SE,
    upper = Estimate + 1.96 * SE,
    sig = ifelse(lower * upper > 0, "yes", "no")  # significant if CI does not cross 0
  )
# Clean variable names
coef_long$Variable <- coef_long$Variable %>%
  str_replace("CoreTranslocated:Depth", "Treatment × Depth") %>%
  str_replace("CoreTranslocated", "Treatment") %>%
  str_replace("Depth", "Depth")
# Set Variable as a factor with desired order
coef_long$Variable <- factor(coef_long$Variable, levels = c("Treatment", "Depth", "Treatment × Depth"))

#family order: alphabetical, Other at bottom ---
family_order_T1 <- c(
  sort(unique(as.character(coef_long$Family[coef_long$Family != "Other"]))),
  "Other"
)
coef_long$Family <- factor(coef_long$Family, levels = rev(family_order_T1))

coef_long$Bay <- factor(coef_long$Bay, levels = c("heated", "control"))

p_coef_final <- ggplot(coef_long, aes(x = Estimate, y = Family, colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper),
                 height = 0.2,
                 position = position_dodge(width = 0.5)) +
  geom_point(shape = 4, size = 2, position = position_dodge(width = 0.5)) +  # Cross
  facet_grid(Bay ~ Variable, space = "free") +
  scale_colour_manual(values = c("yes" = "black", "no" = "grey70")) +
  labs(x = "Estimate", y = NULL, colour = "Significance") +
  theme_bw() +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 9),
    plot.title = element_text(size = 12),
    strip.text = element_text(size = 11),
    panel.border = element_rect(linewidth = 1.25),
    strip.background = element_rect(fill = "white", colour = NA,linewidth = 0.5),
    legend.position = "none",
    panel.spacing = unit(1, "lines"),
    strip.text.y = element_blank(),      # removes "control"/"heated" labels
    strip.background.y = element_blank())

p_coef_final

ggsave("p_coef_final_by_bay_variable_T1.pdf", plot = p_coef_final, width = 22, height = 16, units = "cm", dpi = 1200)

p_coef_T0 <- p_coef_T0 +
  theme(axis.title.x = element_blank())

p_coef_all <- ggarrange(
  p_coef_T0,  
  p_coef_final,  
  labels = c("a", "b"), 
  font.label = list(size = 12),
  nrow = 2,
  align = "hv",
  heights = c(1.09, 2.1)   
)

p_coef_all
ggsave("p_coef_all.pdf", plot = p_coef_all, width = 22, height = 25, units = "cm")

```


