---
title: "PCA for all tube chemical"
author: "Songjun, edited by Romana"
date: "2026-03-05"
output: html_document
---


load package
```{r library package}
library(readxl)
library(ggplot2)
library(grid)
library(patchwork)
library(gridExtra)
library(corrplot)
library(lme4)
library(lmerTest)
library(tidyverse)
library(ggeffects)
library(ggpubr)
library(car)
library(effectsize)
library(DHARMa)

```


#sepearted into T0 dataset and T1 dataset

```{r data}

tube_total<-read_xlsx("combined_chemical_tubes_all.xlsx")

tube_total <- tube_total %>%
  mutate(Core_ID = paste(Site, Core, sep = "_"))

tube_total$Bay[tube_total$Bay=="Warm"]<-"Heated"
tube_total$Depth[tube_total$Depth=="0-1"]<-0
tube_total$Depth[tube_total$Depth=="0-2"]<-0
tube_total$Depth[tube_total$Depth=="1-2"]<-1
tube_total$Depth[tube_total$Depth=="8-9"]<-8
tube_total$Depth[tube_total$Depth=="15-16"]<-15
tube_total$Depth[tube_total$Depth=="22-23"]<-22
tube_total$Depth[tube_total$Depth=="29-30"]<-29
tube_total<-tube_total[tube_total$Depth != "BW",]
tube_total$Depth<-as.numeric(tube_total$Depth)

site_list<-c("K","J","L","I","M","N1","N2","N3","E","A","F","H2","C2","C","A2","B")

tube_T0_to_15<-tube_total[tube_total$Time=="T0",]
tube_T0_to_15<-tube_T0_to_15[tube_T0_to_15$Depth<21,]

tube_T1_to_15<-tube_total[tube_total$Time=="T1",]
tube_T1_to_15<-tube_T1_to_15[tube_T1_to_15$Depth<21,]


tube_T1_heated_to_15<-tube_T1_to_15[tube_T1_to_15$Bay=="Heated",]
tube_T1_control_to_15<-tube_T1_to_15[tube_T1_to_15$Bay=="Control",]

```


#correlation matrix test (to check if some variables are correlated)
#maybe not very related to our case here, but interesting
```{r}

tube_total_geochemical<-tube_total[,c(6,8,9,11,12,14,16)]


str(tube_total_geochemical)
tube_total_geochemical<-as.data.frame(tube_total_geochemical)
test<-na.omit(tube_total_geochemical)

cor(test)


```

#calculate mean to replace NA values
#T0
```{r}

#later prcomp can't accept NA values, so replace each NA value with mean
#so4
so4_mean<-mean(na.omit(tube_T0_to_15$`SO4 (mg/L)`))
tube_T0_to_15$`SO4 (mg/L)`<-replace_na(tube_T0_to_15$`SO4 (mg/L)`, so4_mean )
#no3-n
no3_n_mean<-mean(na.omit(tube_T0_to_15$`NO3(-N)`))
tube_T0_to_15$`NO3(-N)`<-replace_na(tube_T0_to_15$`NO3(-N)`,  no3_n_mean)
#no3
no3_mean<-mean(na.omit(tube_T0_to_15$`NO3 (mg/L)`))
tube_T0_to_15$`NO3 (mg/L)`<-replace_na(tube_T0_to_15$`NO3 (mg/L)`,no3_mean)
#Total FE
total_fe_mean<-mean(na.omit(tube_T0_to_15$`Total Fe (μM)`))
tube_T0_to_15$`Total Fe (μM)`<-replace_na(tube_T0_to_15$`Total Fe (μM)`, total_fe_mean)
#FE 2
fe2_mean<-mean(na.omit(tube_T0_to_15$`Fe2 (μM)`))
tube_T0_to_15$`Fe2 (μM)`<-replace_na(tube_T0_to_15$`Fe2 (μM)`,  fe2_mean)
#po4
po4mean<-mean(na.omit(tube_T0_to_15$`PO4 3- (μM)`))
tube_T0_to_15$`PO4 3- (μM)`<-replace_na(tube_T0_to_15$`PO4 3- (μM)`, po4mean)
#OM
OM_mean<-mean(na.omit(tube_T0_to_15$`organic _percentage`))
tube_T0_to_15$`organic _percentage`<-replace_na(tube_T0_to_15$`organic _percentage`,  OM_mean)


```



#start making PCA
#table prepartion and calculating PC for T0
```{r}


#not taking NO3-N for now as it seems a repeat to NO3 and those two are highly related

tube_T0_to_15_geochemical<-tube_T0_to_15[,c(6,9,11,12,14,16)]


PCA_OKG_T0_to_15<-prcomp(tube_T0_to_15_geochemical, scale. = TRUE, center = TRUE)
PCtable_OKG_T0_to_15<-PCA_OKG_T0_to_15$x
tube_T0_to_15_with_PC_1_2<-cbind(tube_T0_to_15,PCtable_OKG_T0_to_15[,1:2])

coefficient_OKG_T0_to_15<-PCA_OKG_T0_to_15$rotation
coefficient_OKG_T0_to_15
#always consider this explanation table with the PCA results, as it links to the specified how each chemical parameters contribute to each PC axis

coefficient_OKG_T0_to_15<-as.data.frame(coefficient_OKG_T0_to_15)
write.csv(coefficient_OKG_T0_to_15,file = "coefficient_to_15_T0.csv",row.names = TRUE)


#get the proportion of variance
summary(PCA_OKG_T0_to_15)

#PC1 48.96%, PC2 20.88%




```



#linear mixed model for T0
```{r}

#pc1

tube_T0_to_15_with_PC_1_2$Depth<-as.numeric(tube_T0_to_15_with_PC_1_2$Depth)
tube_T0_to_15_model_PC1<-lmer(PC1~Bay*Depth+(1|Core_ID), data = tube_T0_to_15_with_PC_1_2)

summary(tube_T0_to_15_model_PC1)
anova(tube_T0_to_15_model_PC1) 
rand(tube_T0_to_15_model_PC1)# check significance of random
eta_squared(tube_T0_to_15_model_PC1) # compute partial eta2 values

plot(tube_T0_to_15_model_PC1)
hist(resid(tube_T0_to_15_model_PC1))
simulateResiduals(fittedModel = tube_T0_to_15_model_PC1, plot=TRUE)# not perfect but ok


# interaction not significant simplify model - exclude interaction
tube_T0_to_15_model_PC1b<-lmer(PC1~Bay+Depth + (1|Core_ID), data = tube_T0_to_15_with_PC_1_2)

summary(tube_T0_to_15_model_PC1b)
anova(tube_T0_to_15_model_PC1b)
rand(tube_T0_to_15_model_PC1b)
eta_squared(tube_T0_to_15_model_PC1b) # compute partial eta2 values


#pc2
tube_T0_to_15_model_PC2<-lmer(PC2~Bay*Depth+ (1 | Core_ID), data = tube_T0_to_15_with_PC_1_2)
summary(tube_T0_to_15_model_PC2)
anova(tube_T0_to_15_model_PC2, type="II")
rand(tube_T0_to_15_model_PC2)
eta_squared(tube_T0_to_15_model_PC2)

plot(tube_T0_to_15_model_PC2)
hist(resid(tube_T0_to_15_model_PC2))
simulateResiduals(fittedModel = tube_T0_to_15_model_PC2, plot=TRUE)#looks good


#simplified to exclude interaction
tube_T0_to_15_model_PC2b<-lmer(PC2~Bay+Depth+ (1 | Core_ID), data = tube_T0_to_15_with_PC_1_2)
summary(tube_T0_to_15_model_PC2b)
anova(tube_T0_to_15_model_PC2b)
rand(tube_T0_to_15_model_PC2b)
eta_squared(tube_T0_to_15_model_PC2b)

```


#visulaize T0 results
#pc predicted values
```{r}

# Shared theme to reuse across all plots
theme_consistent <- theme_bw() +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 13),
    strip.text = element_text(size = 16),
    panel.border = element_rect(linewidth = 1.25),
    strip.background = element_rect(fill = "white", linewidth = 1.25)
  )

pred_tube_T0_to_15_model_PC1b <- ggeffect(tube_T0_to_15_model_PC1b, terms = c("Depth", "Bay"))
pred_tube_T0_to_15_model_PC1b




tube_T0_to_15_pc1_rawdotplot<-ggplot(pred_tube_T0_to_15_model_PC1b, aes(x = x, y = predicted, color = group, fill = group))+
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, colour = NA) +
  geom_point(data = tube_T0_to_15_with_PC_1_2,
             mapping = aes(x = Depth, y = PC1, color = Bay),
             position = position_jitter(width = 0.1), alpha = 0.5,
             inherit.aes = FALSE) +
  scale_colour_manual(values = c("#2166AC","#D6191B"))+  
  scale_fill_manual(values = c("#2166AC", "#D6191B")) +  
  labs(x = "Depth (cm)",
    y = "PC1")+
  theme_consistent+
  theme(legend.position = "none")

tube_T0_to_15_pc1_rawdotplot


pred_tube_T0_to_15_model_PC2b <- ggeffect(tube_T0_to_15_model_PC2b, terms = c("Depth", "Bay"))
pred_tube_T0_to_15_model_PC2b


tube_T0_to_15_pc2_rawdotplot<-ggplot(pred_tube_T0_to_15_model_PC2b, aes(x = x, y = predicted, color = group, fill = group))+
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, colour = NA) +
  geom_point(data = tube_T0_to_15_with_PC_1_2,
             mapping = aes(x = Depth, y = PC2, color = Bay),
             position = position_jitter(width = 0.1), alpha = 0.5,
             inherit.aes = FALSE) +
  scale_colour_manual(values = c("#2166AC","#D6191B"))+  
  scale_fill_manual(values = c("#2166AC", "#D6191B")) +  
  labs(x = "Depth (cm)",
    y = "PC2")+
  theme_consistent+
  theme(legend.position = "none")

tube_T0_to_15_pc2_rawdotplot







```

#pca dot plot
```{r}

# Ensure tube_T1_to_15_with_PC_1_2 is a data frame
tube_T0_to_15_with_PC_1_2 <- as.data.frame(tube_T0_to_15_with_PC_1_2)

# Check if 'Bay' column exists (replace 'Bay' with actual column name if different)
if (!"Bay" %in% colnames(tube_T0_to_15_with_PC_1_2)) {
  stop("Column 'Bay' not found! Replace it with the correct column name.")
}

# Create a data frame for the PCA scores
pca_data <- tube_T0_to_15_with_PC_1_2

# Extract PCA loadings (rotation) for geochemical variables
pca_arrows <- as.data.frame(PCA_OKG_T0_to_15$rotation[, 1:2])
pca_arrows$Variable <- rownames(pca_arrows)  # Add variable names

# Scale arrows for visualization
scale_factor <- 7  # Adjust as needed for arrow length
pca_arrows <- pca_arrows %>%
  mutate(PC1 = PC1 * scale_factor,
         PC2 = PC2 * scale_factor)


# PCA plot with color by 'Bay' and arrows for geochemical variablesand depths added
pca_biplot_plot_with_depth<-ggplot(pca_data, aes(x = PC1, y = PC2, color = Bay)) +
  geom_point(size = 8, alpha = 0.8) +  # Plot PCA scores with color by Bay
  geom_text(data = pca_data, aes(x = PC1, y = PC2, label = Depth),
                 color = "white", size = 5) + # alternatively could show depth as a colour gradient
  geom_segment(data = pca_arrows, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), color = "black") +  # Arrows for variables
  geom_label_repel(data = pca_arrows, aes(x = PC1, y = PC2, label = Variable),
                 color = "black", size = 4, fill = "white", box.padding = 0.5, segment.color = NA) +
  xlab("PC 1 (48.96 %)") +
  ylab("PC 2 (20.88 %)") +
  #ggtitle("PCA Biplot of T0") +
  xlim(-5,9)+
  ylim(-3.5,6)+
    scale_colour_manual(values = c("#2166AC","#D6191B"))+
  theme_consistent+
  theme(legend.position = "none")

pca_biplot_plot_with_depth

# add depths to figure

```

#select T0 figure mix
```{r}

mix1 <- ggarrange(
  tube_T0_to_15_pc1_rawdotplot,  
  tube_T0_to_15_pc2_rawdotplot,  
  labels = c("b", "c"), 
  font.label = list(size = 18),
  common.legend = TRUE, 
  legend = "none", 
  nrow = 2
)

mix1

# Combine the plots into the custom layout using layout_matrix
pca_mix_tube_T0_to_15 <- ggarrange(
  pca_biplot_plot_with_depth,  
  mix1,
  labels = c("a", "b", "c"), 
  font.label = list(size = 18),
  ncol = 2,
  widths = c(1.5, 1)) 

pca_mix_tube_T0_to_15


ggsave("PCA_T0_mix_plot_to_15.pdf", plot = pca_mix_tube_T0_to_15,width = 25, height = 15,units = "cm")

```




#T1 combined
#calculate mean to replace NA values
#T1
```{r}

#later prcomp can't accept NA values, so replace each NA value with mean
#so4
so4_mean<-mean(na.omit(tube_T1_to_15$`SO4 (mg/L)`))
tube_T1_to_15$`SO4 (mg/L)`<-replace_na(tube_T1_to_15$`SO4 (mg/L)`, so4_mean )
#no3-n
no3_n_mean<-mean(na.omit(tube_T1_to_15$`NO3(-N)`))
tube_T1_to_15$`NO3(-N)`<-replace_na(tube_T1_to_15$`NO3(-N)`,  no3_n_mean)
#no3
no3_mean<-mean(na.omit(tube_T1_to_15$`NO3 (mg/L)`))
tube_T1_to_15$`NO3 (mg/L)`<-replace_na(tube_T1_to_15$`NO3 (mg/L)`,no3_mean)
#Total FE
total_fe_mean<-mean(na.omit(tube_T1_to_15$`Total Fe (μM)`))
tube_T1_to_15$`Total Fe (μM)`<-replace_na(tube_T1_to_15$`Total Fe (μM)`, total_fe_mean)
#FE 2
fe2_mean<-mean(na.omit(tube_T1_to_15$`Fe2 (μM)`))
tube_T1_to_15$`Fe2 (μM)`<-replace_na(tube_T1_to_15$`Fe2 (μM)`,  fe2_mean)
#po4
po4mean<-mean(na.omit(tube_T1_to_15$`PO4 3- (μM)`))
tube_T1_to_15$`PO4 3- (μM)`<-replace_na(tube_T1_to_15$`PO4 3- (μM)`, po4mean)
#OM
OM_mean<-mean(na.omit(tube_T1_to_15$`organic _percentage`))
tube_T1_to_15$`organic _percentage`<-replace_na(tube_T1_to_15$`organic _percentage`,  OM_mean)


```



#start making PCA
#table prepartion and calculating PC for T1
```{r}


tube_T1_to_15_geochemical<-tube_T1_to_15[,c(6,9,11,12,14,16)]


PCA_OKG_T1_to_15<-prcomp(tube_T1_to_15_geochemical, scale. = TRUE, center = TRUE)
PCtable_OKG_T1_to_15<-PCA_OKG_T1_to_15$x
tube_T1_to_15_with_PC_1_2<-cbind(tube_T1_to_15,PCtable_OKG_T1_to_15[,1:2])

coefficient_OKG_T1_to_15<-PCA_OKG_T1_to_15$rotation
coefficient_OKG_T1_to_15
#always consider this explanation table with the PCA results, as it links to the specified how each chemical parameters contribute to each PC axis

coefficient_OKG_T1_to_15<-as.data.frame(coefficient_OKG_T1_to_15)
write.csv(coefficient_OKG_T1_to_15,file = "coefficient_to_15_T1.csv",row.names = TRUE)


#get the proportion of variance
summary(PCA_OKG_T1_to_15)

#PC1 50.64%, PC2 16.96%


```



#linear mixed model for T1
```{r}

#pc1
tube_T1_to_15_with_PC_1_2$Depth<-as.numeric(as.character(tube_T1_to_15_with_PC_1_2$Depth))
tube_T1_to_15_model_PC1<-lmer(PC1~Bay*Depth*Core+(1|Core_ID), data = tube_T1_to_15_with_PC_1_2)
summary(tube_T1_to_15_model_PC1)
anova(tube_T1_to_15_model_PC1)
rand(tube_T1_to_15_model_PC1)
eta_squared(tube_T1_to_15_model_PC1)

plot(tube_T1_to_15_model_PC1)
hist(resid(tube_T1_to_15_model_PC1))
simulateResiduals(fittedModel = tube_T1_to_15_model_PC1, plot=TRUE) # not perfect but ok

#results for T1 mixed model
# for the mixed model with sites as random effect, the three way interaction  (bay, depth, core) is significant

#pc2
tube_T1_to_15_model_PC2<-lmer(PC2~Bay*Depth*Core+(1|Core_ID), data = tube_T1_to_15_with_PC_1_2)
summary(tube_T1_to_15_model_PC2)
anova(tube_T1_to_15_model_PC2) # only depth significant
rand(tube_T1_to_15_model_PC2)
eta_squared(tube_T1_to_15_model_PC2)

plot(tube_T1_to_15_model_PC2)
hist(resid(tube_T1_to_15_model_PC2))
simulateResiduals(fittedModel = tube_T1_to_15_model_PC2, plot=TRUE)

#pc2 without interaction
tube_T1_to_15_model_PC2b<-lmer(PC2~Bay+Depth+Core
                               +Bay*Depth
                               + Bay*Core 
                               +Core*Depth
                               +(1|Core_ID), data = tube_T1_to_15_with_PC_1_2)

anova(tube_T1_to_15_model_PC2b)

tube_T1_to_15_model_PC2c<-lmer(PC2~Bay+Depth+Core
                               +(1|Core_ID), data = tube_T1_to_15_with_PC_1_2)

anova(tube_T1_to_15_model_PC2c)

```

#linear mixed model for T1 two bays seperated
```{r}


#pc1

#heated
tube_T1_to_15_model_PC1_heated<-lmer(PC1~Depth*Core+(1|Core_ID), data = tube_T1_to_15_with_PC_1_2[tube_T1_to_15_with_PC_1_2$Bay=="Heated",])

summary(tube_T1_to_15_model_PC1_heated)
anova(tube_T1_to_15_model_PC1_heated)
rand(tube_T1_to_15_model_PC1_heated)

#control
tube_T1_to_15_model_PC1_control<-lmer(PC1~Depth*Core+(1|Core_ID), data = tube_T1_to_15_with_PC_1_2[tube_T1_to_15_with_PC_1_2$Bay=="Control",])

summary(tube_T1_to_15_model_PC1_control)
anova(tube_T1_to_15_model_PC1_control)

rand(tube_T1_to_15_model_PC1_control)


```



#visulaize T1 results

#heated bay
#pc predicted values
```{r}


tube_T1_to_15_with_PC_1_2

tube_T1_to_15_pc1_prediction <- predict_response(tube_T1_to_15_model_PC1, terms = c("Depth", "Core", "Bay"))


tube_T1_heated_to_15_pc1_rawdotplot<-ggplot(tube_T1_to_15_pc1_prediction[tube_T1_to_15_pc1_prediction$facet=="Heated",], aes(x = x, y = predicted, color = group, fill = group)) +
    geom_line(aes(linetype = group), linewidth = 1) + 
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, colour = NA) +
  geom_point(data = tube_T1_to_15_with_PC_1_2[tube_T1_to_15_with_PC_1_2$Bay=="Heated",],
             mapping = aes(x = Depth, y = PC1, color = Core),
             position = position_jitter(width = 0.1), alpha = 0.5,
             inherit.aes = FALSE) +
  scale_colour_manual(values = c("#D6191B","#00ACC1"))+  
  scale_fill_manual(values = c("#D6191B","#00ACC1")) +  
      scale_linetype_manual(values = c("solid", "dashed"),
                        guide = "none") + 
  ylim(-3,6)+
  labs(x = "Depth (cm)",
    y = "PC1")+
  ggtitle("Heated bay")+
  theme_consistent

tube_T1_heated_to_15_pc1_rawdotplot


tube_T1_control_to_15_pc1_rawdotplot<-ggplot(tube_T1_to_15_pc1_prediction[tube_T1_to_15_pc1_prediction$facet=="Control",], aes(x = x, y = predicted, color = group, fill = group)) +
    geom_line(aes(linetype = group), linewidth = 1) +  
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, colour = NA) +
  geom_point(data = tube_T1_to_15_with_PC_1_2[tube_T1_to_15_with_PC_1_2$Bay=="Control",],
             mapping = aes(x = Depth, y = PC1, color = Core),
             position = position_jitter(width = 0.1), alpha = 0.5,
             inherit.aes = FALSE) +
  scale_colour_manual(values = c("#2166AC","#F57C00"))+  
  scale_fill_manual(values = c("#2166AC","#F57C00")) +  
    scale_linetype_manual(values = c("solid", "dashed"),
                        guide = "none") + 
  ylim(-3,6)+
  labs(x = "Depth (cm)",
    y = "PC1")+
  ggtitle("Control bay")+
  theme_consistent

tube_T1_control_to_15_pc1_rawdotplot



```

#pca dot plot heated bay
```{r}

# Ensure tube_T1_to_15_with_PC_1_2 is a data frame
tube_T1_to_15_with_PC_1_2 <- as.data.frame(tube_T1_to_15_with_PC_1_2)

# Check if 'Bay' column exists (replace 'Bay' with actual column name if different)
if (!"Bay" %in% colnames(tube_T0_to_15_with_PC_1_2)) {
  stop("Column 'Bay' not found! Replace it with the correct column name.")
}


# Extract PCA loadings (rotation) for geochemical variables
pca_arrows<- as.data.frame(PCA_OKG_T1_to_15$rotation[, 1:2])
pca_arrows$Variable <- rownames(pca_arrows)  # Add variable names

# Scale arrows for visualization
scale_factor <- 5.5  # Adjust as needed for arrow length
pca_arrows<- pca_arrows%>%
  mutate(PC1 = PC1 * scale_factor,
         PC2 = PC2 * scale_factor)


# Create a data frame for the PCA scores heated bay
pca_data_heated <- tube_T1_to_15_with_PC_1_2[tube_T1_to_15_with_PC_1_2$Bay=="Heated",]


# heated bay pca biplot 
pca_biplot_plot_with_depth_heated<-ggplot(pca_data_heated, aes(x = PC1, y = PC2, color = Core)) +
  geom_point(size = 8, alpha = 0.8) +  # Plot PCA scores with color by Bay
  geom_text(data = pca_data_heated, aes(x = PC1, y = PC2, label = Depth),
                 color = "white", size = 5) + # alternatively could show depth as a colour gradient
  geom_segment(data = pca_arrows, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), color = "black") +  # Arrows for variables
  geom_label_repel(data = pca_arrows, aes(x = PC1, y = PC2, label = Variable),
                 color = "black", size = 4, fill = "white", box.padding = 0.5, segment.color = NA) +
  xlab("PC 1 (50.64 %)") +
  ylab("PC 2 (16.96 %)") +
  ggtitle("Heated bay") +
  xlim(-3,6.5)+
  ylim(-6.5,3)+
    scale_colour_manual(values = c("#D6191B","#00ACC1"))+
  theme_consistent +
  theme(legend.position = "none")
pca_biplot_plot_with_depth_heated


# Create a data frame for the PCA scores control bay
pca_data_control <- tube_T1_to_15_with_PC_1_2[tube_T1_to_15_with_PC_1_2$Bay=="Control",]


# or with labels instead and depths added? 
pca_biplot_plot_with_depth_control<-ggplot(pca_data_control, aes(x = PC1, y = PC2, color = Core)) +
  geom_point(size = 8, alpha = 0.8) +  # Plot PCA scores with color by Bay
  geom_text(data = pca_data_control, aes(x = PC1, y = PC2, label = Depth),
                 color = "white", size = 5) + # alternatively could show depth as a colour gradient
  geom_segment(data = pca_arrows, aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), color = "black") +  # Arrows for variables
  geom_label_repel(data = pca_arrows, aes(x = PC1, y = PC2, label = Variable),
                 color = "black", size = 4, fill = "white", box.padding = 0.5, segment.color = NA) +
  xlab("PC 1 (50.64 %)") +
  ylab("PC 2 (16.96 %)") +
  ggtitle("Control bay") +
   xlim(-3,6.5)+
  ylim(-6.5,3)+
    scale_colour_manual(values = c("#2166AC","#F57C00"))+
  theme_consistent +
  theme(legend.position = "none")

  
pca_biplot_plot_with_depth_control


```

#select T1 figure mix
```{r}


# Combine the plots into the custom layout using layout_matrix
pca_mix_tube_T1_to_15 <- ggarrange(
  pca_biplot_plot_with_depth_heated,  
  tube_T1_heated_to_15_pc1_rawdotplot,
  pca_biplot_plot_with_depth_control,
  tube_T1_control_to_15_pc1_rawdotplot,
  labels = c("a", "b", "c","d"), 
  font.label = list(size = 18),
  align = "hv",
  common.legend = TRUE, legend = "right",
                    ncol=2, nrow=2)

pca_mix_tube_T1_to_15


ggsave("PCA_T1_mix_plot_to_15.pdf", plot = pca_mix_tube_T1_to_15, width = 30, height = 30,units = "cm")



```















